---
title: "Regression Model for Superconductivity Temperature Using R Take 3"
author: "David Lowe"
date: "July 22, 2019"
output: 
  html_document: 
    toc: yes
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Template Credit: Adapted from a template made available by Dr. Jason Brownlee of Machine Learning Mastery. [http://machinelearningmastery.com/]

SUMMARY: The purpose of this project is to construct a prediction model using various machine learning algorithms and to document the end-to-end steps using a template. The Superconductivity Temperature dataset is a regression situation where we are trying to predict the value of a continuous variable.

INTRODUCTION: The research team wishes to create a statistical model for predicting the superconducting critical temperature based on the features extracted from the superconductor’s chemical formula. The model seeks to examine the features that can contribute the most to the model’s predictive accuracy.

In iteration Take1, we established the baseline mean squared error for comparison with the future rounds of modeling.

In iteration Take2, we examined the feature selection technique of attribute importance ranking by using the Gradient Boosting algorithm. By selecting only the most important attributes, we decreased the modeling time and still maintained a similar level of RMSE compared to the baseline model.

In this iteration, we will examine the feature selection technique of recursive feature elimination (RFE) by using the Bagged Trees algorithm. By selecting no more than 50 attributes, we hope to maintain a similar level of RMSE compared to the baseline model.

ANALYSIS: The baseline performance of the machine learning algorithms achieved an average RMSE of 16.33. Two algorithms (Random Forest and Gradient Boosting) achieved the top RMSE metrics after the first round of modeling. After a series of tuning trials, Random Forest turned in the best overall result and achieved an RMSE metric of 9.72. By using the optimized parameters, the Random Forest algorithm processed the test dataset with an RMSE of 9.40, which was even better than the prediction from the training data.

From iteration Take2, the average performance of the machine learning algorithms achieved an RMSE of 16.40. Random Forest achieved an RMSE metric of 9.73 with the training dataset and processed the test dataset with an RMSE of 9.39. At the importance level of 99%, the attribute importance technique eliminated 10 of 81 total attributes. The remaining 71 attributes produced a model that achieved a comparable RMSE to the baseline model. The modeling time went from 6 hours 26 minutes down to 5 hours 50 minutes, a saving of 9.3%.

From iteration Take3, the average performance of the machine learning algorithms achieved an RMSE of 16.63. Random Forest achieved an RMSE metric of 9.77 with the training dataset and processed the test dataset with an RMSE of 9.43. At the importance level of 99%, the attribute importance technique eliminated 35 of 81 total attributes. The remaining 46 attributes produced a model that achieved a comparable RMSE to the baseline model. The modeling time went from 6 hours 26 minutes down to 3 hours 51 minutes, a saving of 40.1%.

CONCLUSION: For this iteration, the Random Forest algorithm achieved the best overall results using the training and testing datasets. For this dataset, Random Forest should be considered for further modeling.

Dataset Used: Superconductivity Data Data Set

Dataset ML Model: Regression with numerical attributes

Dataset Reference: https://archive.ics.uci.edu/ml/datasets/Superconductivty+Data

One potential source of performance benchmarks: https://doi.org/10.1016/j.commatsci.2018.07.052

The project aims to touch on the following areas:

1. Document a predictive modeling problem end-to-end.
2. Explore data cleaning and transformation options
3. Explore non-ensemble and ensemble algorithms for baseline model performance
4. Explore algorithm tuning techniques for improving model performance

Any predictive modeling machine learning project genrally can be broken down into about six major tasks:

1. Prepare Problem
2. Summarize Data
3. Prepare Data
4. Model and Evaluate Algorithms
5. Improve Accuracy or Results
6. Finalize Model and Present Results

## 1. Prepare Problem

### 1.a) Load libraries

```{r}
startTimeScript <- proc.time()
library(caret)
library(corrplot)
library(DMwR)
library(Hmisc)
library(mailR)
library(ROCR)
library(stringr)
```

```{r}
# Create one random seed number for reproducible results
seedNum <- 888
set.seed(seedNum)

# Set up the flag to send status emails (setting TRUE will send!)
notifyStatus <- TRUE
```

### 1.b) Set up the email notification function

```{r}
email_notify <- function(msg=""){
  sender <- Sys.getenv("MAIL_SENDER")
  receiver <- Sys.getenv("MAIL_RECEIVER")
  gateway <- Sys.getenv("SMTP_GATEWAY")
  smtpuser <- Sys.getenv("SMTP_USERNAME")
  password <- Sys.getenv("SMTP_PASSWORD")
  sbj_line <- "Notification from R Binary Classification Script"
  send.mail(
    from = sender,
    to = receiver,
    subject= sbj_line,
    body = msg,
    smtp = list(host.name = gateway, port = 587, user.name = smtpuser, passwd = password, ssl = TRUE),
    authenticate = TRUE,
    send = TRUE)
}
```

```{r}
if (notifyStatus) email_notify(paste("Library and Data Loading has begun!",date()))
```

### 1.c) Load dataset

```{r}
# Slicing up the document path to get the final destination file name
dataset_path <- 'https://archive.ics.uci.edu/ml/machine-learning-databases/00464/superconduct.zip'
doc_path_list <- str_split(dataset_path, "/")
dest_file <- doc_path_list[[1]][length(doc_path_list[[1]])]

if (!file.exists(dest_file)) {
  # Download the document from the website
  cat("Downloading", dataset_path, "as", dest_file, "\n")
  download.file(dataset_path, dest_file, mode = "wb")
  cat(dest_file, "downloaded!\n")
  unzip(dest_file)
  cat(dest_file, "unpacked!\n")
}

inputFile <- "train.csv"
xy_original <- read.csv(inputFile, header = TRUE, sep = ",")
```

```{r}
# Take a peek at the dataframe after the import
head(xy_original)
```

```{r}
sapply(xy_original, class)
```

```{r}
sapply(xy_original, function(x) sum(is.na(x)))
```

### 1.d) Data Cleaning

```{r}
# Not applicable for this iteration of modeling
```

### 1.e) Splitting Data into Training and Testing Sets

```{r}
# Use variable totCol to hold the number of columns in the dataframe
totCol <- ncol(xy_original)

# Set up variable totAttr for the total number of attribute columns
totAttr <- totCol-1
```

```{r}
# targetCol variable indicates the column location of the target/class variable
# If the first column, set targetCol to 1. If the last column, set targetCol to totCol
# if (targetCol <> 1) and (targetCol <> totCol), be aware when slicing up the dataframes for visualization! 
targetCol <- totCol

# Standardize the class column to the name of targetVar if applicable
colnames(xy_original)[targetCol] <- "targetVar"
```

```{r}
# We create various datasets (x_original, y_original) for various visualization and cleaning/transformation operations.
if (targetCol==1) {
x_original <- xy_original[,(targetCol+1):totCol]
y_original <- xy_original[,targetCol]
} else {
x_original <- xy_original[,1:(totAttr)]
y_original <- xy_original[,totCol]
}
```

### 1.f) Set up the key parameters to be used in the script

```{r}
# Set up the number of row and columns for visualization display. dispRow * dispCol should be >= totAttr
dispCol <- 4
if (totAttr%%dispCol == 0) {
dispRow <- totAttr%/%dispCol
} else {
dispRow <- (totAttr%/%dispCol) + 1
}
cat("Will attempt to create graphics grid (col x row): ", dispCol, ' by ', dispRow)
```

```{r}
# Run algorithms using 10-fold cross validation
control <- trainControl(method="repeatedcv", number=10, repeats=1)
metricTarget <- "RMSE"
```

```{r}
if (notifyStatus) email_notify(paste("Library and Data Loading completed!",date()))
```

## 2. Summarize Data
To gain a better understanding of the data that we have on-hand, we will leverage a number of descriptive statistics and data visualization techniques. The plan is to use the results to consider new questions, review assumptions, and validate hypotheses that we can investigate later with specialized models.

```{r}
if (notifyStatus) email_notify(paste("Data Summarization and Visualization has begun!",date()))
```

### 2.a) Descriptive statistics

#### 2.a.i) Peek at the data itself.

```{r}
head(xy_original)
```

#### 2.a.ii) Dimensions of the dataset.

```{r}
dim(xy_original)
```

#### 2.a.iii) Types of the attributes.

```{r}
sapply(xy_original, class)
```

#### 2.a.iv) Statistical summary of all attributes.

```{r}
summary(xy_original)
```

#### 2.a.v) Count missing values.

```{r}
sapply(xy_original, function(x) sum(is.na(x)))
```

### 2.b) Data visualizations

```{r}
# Boxplots for each attribute
# par(mfrow=c(dispRow,dispCol))
# for(i in 1:totAttr) {
# 	boxplot(x_original[,i], main=names(x_original)[i])
# }
```

```{r}
# Histograms each attribute
# par(mfrow=c(dispRow,dispCol))
# for(i in 1:totAttr) {
# 	hist(x_original[,i], main=names(x_original)[i])
# }
```

```{r}
# Density plot for each attribute
# par(mfrow=c(dispRow,dispCol))
# for(i in 1:totAttr) {
# 	plot(density(x_original[,i]), main=names(x_original)[i])
# }
```

```{r}
# Correlation plot
# correlations <- cor(x_original)
# corrplot(correlations, method="circle")
```

```{r}
if (notifyStatus) email_notify(paste("Data Summarization and Visualization completed!",date()))
```

## 3. Prepare Data
Some dataset may require additional preparation activities that will best exposes the structure of the problem and the relationships between the input attributes and the output variable. Some data-prep tasks might include:

* Cleaning data by removing duplicates, marking missing values and even imputing missing values.
* Feature selection where redundant features may be removed.
* Data transforms where attributes are scaled or redistributed in order to best expose the structure of the problem later to learning algorithms.

```{r}
if (notifyStatus) email_notify(paste("Data Cleaning and Transformation has begun!",date()))
```

### 3.a) Split Original Dataset into Training and Test

```{r}
# Create a list of the rows in the original dataset we can use for training
# Use 70% of the data to train the models and the remaining for testing/validation
set.seed(seedNum)
training_index <- createDataPartition(xy_original$targetVar, p=0.70, list=FALSE)
xy_train <- xy_original[training_index,]
xy_test <- xy_original[-training_index,]

totCol <- ncol(xy_original)
targetCol <- totCol
if (targetCol==1) {
y_test <- xy_test[,targetCol]
} else {
y_test <- xy_test[,totCol]
}
```

### 3.b) Feature Selection

```{r}
# Sample code for perfoming feature selection using the Recursive Feature Elimination (RFE) technique
startTimeModule <- proc.time()
set.seed(seedNum)
x_rfe <- xy_train[,1:totAttr]
y_rfe <- xy_train[,totCol]
normalization <- preProcess(x_rfe)
x_rfe <- predict(normalization, x_rfe)
x_rfe <- as.data.frame(x_rfe)
rfeCTRL <- rfeControl(functions=treebagFuncs, method="cv", number=10, repeats=1, verbose=FALSE, returnResamp="all")
optimalVars <- 50
subsets <- c(2:optimalVars)
rfeProfile <- rfe(x_rfe, y_rfe, sizes=subsets, rfeControl=rfeCTRL)
print(rfeProfile)
plot(rfeProfile, type=c("g", "o"))
```

```{r}
numberRFEVars <- length(predictors(rfeProfile))
if (numberRFEVars <= optimalVars) {
  rfeAttributes <- predictors(rfeProfile)
} else {
  newProfile <- update(rfeProfile, x=x_rfe, y=y_rfe, size=optimalVars)
  rfeAttributes <- newProfile$bestVar
}
cat('Number of attributes selected from the RFE algorithm:',length(rfeAttributes),'\n')
print(rfeAttributes)
```

```{r}
# Removing the unselected attributes from the training and validation dataframes
rfeAttributes <- c(rfeAttributes,"targetVar")
xy_train <- xy_train[, (names(xy_train) %in% rfeAttributes)]
xy_test <- xy_test[, (names(xy_test) %in% rfeAttributes)]
```

### 3.c) Data Transforms

```{r}
# Not applicable for this iteration of modeling
```

### 3.d) Display the Final Dataset for Model-Building

```{r}
dim(xy_train)
dim(xy_test)
```

```{r}
sapply(xy_train, class)
```

```{r}
if (notifyStatus) email_notify(paste("Data Cleaning and Transformation completed!",date()))
proc.time()-startTimeScript
```

## 4. Model and Evaluate Algorithms
After the data-prep, we next work on finding a workable model by evaluating a subset of machine learning algorithms that are good at exploiting the structure of the dataset. The typical evaluation tasks include:

* Defining test options such as cross validation and the evaluation metric to use.
* Spot checking a suite of linear and nonlinear machine learning algorithms.
* Comparing the estimated accuracy of algorithms.

For this project, we will evaluate four linear, one non-linear, and three ensemble algorithms:

Linear Algorithm: Linear Regression, Ridge, The Lasso, Elasticnet

Non-Linear Algorithm: Decision Trees (CART)

Ensemble Algorithms: Bagged CART, Random Forest, and Gradient Boosting

The random number seed is reset before each run to ensure that the evaluation of each algorithm is performed using the same data splits. It ensures the results are directly comparable.

### 4.a) Generate models using linear algorithms

```{r}
startModeling <- proc.time()
```

```{r LR}
# Linear Regression (Regression)
if (notifyStatus) email_notify(paste("Linear Regression modeling has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
fit.lm <- train(targetVar~., data=xy_train, method="lm", metric=metricTarget, trControl=control)
print(fit.lm)
proc.time()-startTimeModule
if (notifyStatus) email_notify(paste("Linear Regression modeling completed!",date()))
```

```{r RIDGE}
# Ridge (Regression)
if (notifyStatus) email_notify(paste("Ridge Regression modeling has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
fit.ridge <- train(targetVar~., data=xy_train, method="ridge", metric=metricTarget, trControl=control)
print(fit.ridge)
proc.time()-startTimeModule
if (notifyStatus) email_notify(paste("Ridge Regression modeling completed!",date()))
```

```{r LASSO}
# The Lasso (Regression)
if (notifyStatus) email_notify(paste("The Lasso modeling has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
fit.lasso <- train(targetVar~., data=xy_train, method="lasso", metric=metricTarget, trControl=control)
print(fit.lasso)
proc.time()-startTimeModule
if (notifyStatus) email_notify(paste("The Lasso modeling completed!",date()))
```

```{r ENET}
# Elasticnet (Regression)
if (notifyStatus) email_notify(paste("Elasticnet modeling has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
fit.enet <- train(targetVar~., data=xy_train, method="enet", metric=metricTarget, trControl=control)
print(fit.enet)
proc.time()-startTimeModule
if (notifyStatus) email_notify(paste("Elasticnet modeling completed!",date()))
```

### 4.b) Generate models using nonlinear algorithms

```{r CART}
# Decision Tree - CART (Regression/Classification)
if (notifyStatus) email_notify(paste("Decision Tree modeling has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
fit.cart <- train(targetVar~., data=xy_train, method="rpart", metric=metricTarget, trControl=control)
print(fit.cart)
proc.time()-startTimeModule
if (notifyStatus) email_notify(paste("Decision Tree modeling completed!",date()))
```

### 4.c) Generate models using ensemble algorithms
In this section, we will explore the use and tuning of ensemble algorithms to see whether we can improve the results.

```{r BAGCART}
# Bagged CART (Regression/Classification)
if (notifyStatus) email_notify(paste("Bagged CART modeling has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
fit.bagcart <- train(targetVar~., data=xy_train, method="treebag", metric=metricTarget, trControl=control)
print(fit.bagcart)
proc.time()-startTimeModule
if (notifyStatus) email_notify(paste("Bagged CART modeling completed!",date()))
```

```{r RF}
# Random Forest (Regression/Classification)
if (notifyStatus) email_notify(paste("Random Forest modeling has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
fit.rf <- train(targetVar~., data=xy_train, method="rf", metric=metricTarget, trControl=control)
print(fit.rf)
proc.time()-startTimeModule
if (notifyStatus) email_notify(paste("Random Forest modeling completed!",date()))
```

```{r GBM}
# Gradient Boosting (Regression/Classification)
if (notifyStatus) email_notify(paste("Gradient Boosting modeling has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
# fit.gbm <- train(targetVar~., data=xy_train, method="xgbTree", metric=metricTarget, trControl=control, verbose=F)
fit.gbm <- train(targetVar~., data=xy_train, method="gbm", metric=metricTarget, trControl=control, verbose=F)
print(fit.gbm)
proc.time()-startTimeModule
if (notifyStatus) email_notify(paste("Gradient Boosting modeling completed!",date()))
```

### 4.d) Compare baseline algorithms

```{r SPOT_CHECK}
results <- resamples(list(LR=fit.lm, RIDGE=fit.ridge, LASSO=fit.lasso, ENET=fit.enet, CART=fit.cart, BagCART=fit.bagcart, RF=fit.rf, GBM=fit.gbm))
summary(results)
dotplot(results)
cat('The average RMSE from all models is:',
    mean(c(results$values$`LR~RMSE`,results$values$`RIDGE~RMSE`,results$values$`LASSO~RMSE`,results$values$`ENET~RMSE`,results$values$`CART~RMSE`,results$values$`BagCART~RMSE`,results$values$`RF~RMSE`,results$values$`GBM~RMSE`)),'\n')
cat('Total training time for all models:',proc.time()-startModeling)
```

## 5. Improve Accuracy or Results
After we achieve a short list of machine learning algorithms with good level of accuracy, we can leverage ways to improve the accuracy of the models.

Using the three best-perfoming algorithms from the previous section, we will Search for a combination of parameters for each algorithm that yields the best results.

### 5.a) Algorithm Tuning
Finally, we will tune the best-performing algorithms from each group further and see whether we can get more accuracy out of them.

```{r FINAL1}
# Tuning algorithm #1 - Random Forest
if (notifyStatus) email_notify(paste("Algorithm #1 tuning has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
grid <- expand.grid(mtry = c(10,19,28,37,46))
fit.final1 <- train(targetVar~., data=xy_train, method="rf", metric=metricTarget, tuneGrid=grid, trControl=control)
plot(fit.final1)
print(fit.final1)
proc.time()-startTimeModule
if (notifyStatus) email_notify(paste("Algorithm #1 tuning completed!",date()))
```

```{r FINAL2}
# Tuning algorithm #2 - Gradient Boosting
if (notifyStatus) email_notify(paste("Algorithm #2 tuning has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
grid <- expand.grid(n.trees = c(100,200,300,400,500), interaction.depth = 3, shrinkage = 0.1, n.minobsinnode = 10)
fit.final2 <- train(targetVar~., data=xy_train, method="gbm", metric=metricTarget, tuneGrid=grid, trControl=control, verbose=F)
plot(fit.final2)
print(fit.final2)
proc.time()-startTimeModule
if (notifyStatus) email_notify(paste("Algorithm #2 tuning completed!",date()))
```

### 5.d) Compare Algorithms After Tuning

```{r POST_TUNING}
results <- resamples(list(RF=fit.final1, GBM=fit.final2))
summary(results)
dotplot(results)
```

## 6. Finalize Model and Present Results
Once we have narrow down to a model that we believe can make accurate predictions on unseen data, we are ready to finalize it. Finalizing a model may involve sub-tasks such as:

* Using an optimal model tuned to make predictions on unseen data.
* Creating a standalone model using the tuned parameters
* Saving an optimal model to file for later use.

```{r}
if (notifyStatus) email_notify(paste("Model Validation and Final Model Creation has begun!",date()))
```

### 6.a) Predictions on validation dataset

```{r PREDICT1}
predictions <- predict(fit.final1, newdata=xy_test)
print(RMSE(predictions, y_test))
print(R2(predictions, y_test))
```

```{r PREDICT2}
predictions <- predict(fit.final2, newdata=xy_test)
print(RMSE(predictions, y_test))
print(R2(predictions, y_test))
```

### 6.b) Create standalone model on entire training dataset

```{r FINALMODEL}
startTimeModule <- proc.time()
library(randomForest)
set.seed(seedNum)

# Combining the training and test datasets to form the original dataset that will be used for training the final model
xy_complete <- rbind(xy_train, xy_test)

# finalModel <- randomForest(targetVar~., xy_complete, mtry=28)
# summary(finalModel)
proc.time()-startTimeModule
```

### 6.c) Save model for later use

```{r}
#saveRDS(finalModel, "./finalModel_Regression.rds")
```

```{r}
if (notifyStatus) email_notify(paste("Model Validation and Final Model Creation Completed!",date()))
proc.time()-startTimeScript
```
