---
title: "Multi-Class Classification Model for [Project Name] Using R version 06"
author: "David Lowe"
date: "December 10, 2018"
output: 
  html_document: 
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Template Credit: Adapted from a template made available by Dr. Jason Brownlee of Machine Learning Mastery. [http://machinelearningmastery.com/]

SUMMARY: [The purpose of this project is to construct a prediction model using various machine learning algorithms and to document the end-to-end steps using a template. The [Dataset Name] is a multi-class classification situation where we are trying to predict one of the several possible outcomes.]

INTRODUCTION: [Sample Paragraph - The dataset involves predicting, from USA Forensic Science Service, six types of glass; defined in terms of their oxide content (i.e. Na, Fe, K, etc). The study of classification of types of glass was also partly motivated by criminological investigation. At the scene of the crime, the glass left can be used as evidence...if it is correctly identified!]

ANALYSIS: [Sample Paragraph - The baseline performance of the eight algorithms achieved an average accuracy of 71.52%. Three algorithms (Bagged CART, Random Forest, and Stochastic Gradient Boosting) achieved the top three accuracy scores after the first round of modeling. After a series of tuning trials, Random Forest turned in the top result using the training data. It achieved an average accuracy of 78.04%. Using the optimized tuning parameter available, the Random Forest algorithm processed the validation dataset with an accuracy of 78.69%, which was even better the accuracy of the training data.]

CONCLUSION: [Sample Paragraph - For this iteration, the Random Forest algorithm achieved the top-tier training and validation results. For this dataset, Random Forest should be considered for further modeling or production use.]

Dataset Used: [Dataset Name - Glass Identification Dataset]

Dataset ML Model: Multi-class classification with [numerical | categorical] attributes

Dataset Reference: [Dataset URL - https://archive.ics.uci.edu/ml/datasets/glass+identification]

One potential source of performance benchmarks: [Benchmark URL - https://www.kaggle.com/uciml/glass]

The project aims to touch on the following areas:

1. Document a predictive modeling problem end-to-end.
2. Explore data cleaning and transformation options
3. Explore non-ensemble and ensemble algorithms for baseline model performance
4. Explore algorithm tuning techniques for improving model performance

Any predictive modeling machine learning project genrally can be broken down into about six major tasks:

1. Prepare Problem
2. Summarize Data
3. Prepare Data
4. Model and Evaluate Algorithms
5. Improve Accuracy or Results
6. Finalize Model and Present Results

## 1. Prepare Problem

### 1.a) Load libraries

```{r}
startTimeScript <- proc.time()
library(caret)
library(corrplot)
library(mailR)
library(parallel)
library(mlbench)

# Create one random seed number for reproducible results
seedNum <- 888
set.seed(seedNum)
```

### 1.b) Set up the email notification function

```{r}
email_notify <- function(msg=""){
  sender <- "luozhi2488@gmail.com"
  receiver <- "dave@contactdavidlowe.com"
  sbj_line <- "Notification from R Script"
  password <- readLines("email_credential.txt")
  send.mail(
    from = sender,
    to = receiver,
    subject= sbj_line,
    body = msg,
    smtp = list(host.name = "smtp.gmail.com", port = 465, user.name = sender, passwd = password, ssl = TRUE),
    authenticate = TRUE,
    send = TRUE)
}
```

```{r}
email_notify(paste("Library and Data Loading has begun!",date()))
```

### 1.c) Load dataset

```{r}
#originalDataset <- read.csv("some_dataset.csv", header= TRUE, stringsAsFactors = FALSE)
data(Glass)
originalDataset <- Glass

# Different ways of reading and processing the input dataset. Saving these for future references.
#x_train <- read.fwf("X_train.txt", widths = widthVector, col.names = colNames)
#y_train <- read.csv("y_train.txt", header = FALSE, col.names = c("targetVar"))
#y_train$targetVar <- as.factor(y_train$targetVar)
#xy_train <- cbind(x_train, y_train)
```


```{r}
# Use variable totCol to hold the number of columns in the dataframe
totCol <- ncol(originalDataset)

# Set up variable totAttr for the total number of attribute columns
totAttr <- totCol-1
```

```{r}
# targetCol variable indicates the column location of the target/class variable
# If the first column, set targetCol to 1. If the last column, set targetCol to totCol
# if (targetCol <> 1) and (targetCol <> totCol), be aware when slicing up the dataframes for visualization! 
targetCol <- totCol
colnames(originalDataset)[targetCol] <- "targetVar"
```

```{r}
# We create training datasets (xy_train, x_train, y_train) for various operations.
# We create validation datasets (xy_test, x_test, y_test) for various operations.
set.seed(seedNum)

# Create a list of the rows in the original dataset we can use for training
training_index <- createDataPartition(originalDataset$targetVar, p=0.70, list=FALSE)
# Use 70% of the data to train the models and the remaining for testing/validation
xy_train <- originalDataset[training_index,]
xy_test <- originalDataset[-training_index,]

if (targetCol==1) {
x_train <- xy_train[,(targetCol+1):totCol]
y_train <- xy_train[,targetCol]
y_test <- xy_test[,targetCol]
} else {
x_train <- xy_train[,1:(totAttr)]
y_train <- xy_train[,totCol]
y_test <- xy_test[,totCol]
}
```

### 1.d) Set up the key parameters to be used in the script

```{r}
# Set up the number of row and columns for visualization display. dispRow * dispCol should be >= totAttr
dispCol <- 3
if (totAttr%%dispCol == 0) {
dispRow <- totAttr%/%dispCol
} else {
dispRow <- (totAttr%/%dispCol) + 1
}
cat("Will attempt to create graphics grid (col x row): ", dispCol, ' by ', dispRow)
```

### 1.e) Set test options and evaluation metric

```{r}
# Run algorithms using 10-fold cross validation
control <- trainControl(method="repeatedcv", number=10, repeats=1)
metricTarget <- "Accuracy"
```

```{r}
email_notify(paste("Library and Data Loading completed!",date()))
```

## 2. Summarize Data
To gain a better understanding of the data that we have on-hand, we will leverage a number of descriptive statistics and data visualization techniques. The plan is to use the results to consider new questions, review assumptions, and validate hypotheses that we can investigate later with specialized models.

```{r}
email_notify(paste("Data Summary and Visualization has begun!",date()))
```

### 2.a) Descriptive statistics

#### 2.a.i) Peek at the data itself.

```{r}
head(xy_train)
```

#### 2.a.ii) Dimensions of the dataset.

```{r}
dim(xy_train)
```

#### 2.a.iii) Types of the attributes.

```{r}
sapply(xy_train, class)
```

#### 2.a.iv) Statistical summary of all attributes.

```{r}
summary(xy_train)
```

#### 2.a.v) Summarize the levels of the class attribute.

```{r}
cbind(freq=table(y_train), percentage=prop.table(table(y_train))*100)
```

#### 2.a.vi) Count missing values.

```{r}
sapply(xy_train, function(x) sum(is.na(x)))
```

### 2.b) Data visualizations

#### 2.b.i) Univariate plots to better understand each attribute.

```{r}
# Boxplots for each attribute
# par(mfrow=c(dispRow,dispCol))
for(i in 1:totAttr) {
	boxplot(x_train[,i], main=names(x_train)[i])
}
```

```{r}
# Histograms each attribute
# par(mfrow=c(dispRow,dispCol))
for(i in 1:totAttr) {
	hist(x_train[,i], main=names(x_train)[i])
}
```

```{r}
# Density plot for each attribute
# par(mfrow=c(dispRow,dispCol))
for(i in 1:totAttr) {
	plot(density(x_train[,i]), main=names(x_train)[i])
}
```

#### 2.b.ii) Multivariate plots to better understand the relationships between attributes

```{r}
# Scatterplot matrix colored by class
pairs(targetVar~., data=xy_train, col=xy_train$targetVar)
```

```{r}
# Box and whisker plots for each attribute by class
scales <- list(x=list(relation="free"), y=list(relation="free"))
featurePlot(x=x_train, y=y_train, plot="box", scales=scales)
```

```{r}
# Density plots for each attribute by class value
featurePlot(x=x_train, y=y_train, plot="density", scales=scales)
```

```{r}
# Correlation plot
correlations <- cor(x_train)
corrplot(correlations, method="circle")
```

```{r}
email_notify(paste("Data Summary and Visualization completed!",date()))
```

## 3. Prepare Data
Some dataset may require additional preparation activities that will best exposes the structure of the problem and the relationships between the input attributes and the output variable. Some data-prep tasks might include:

* Cleaning data by removing duplicates, marking missing values and even imputing missing values.
* Feature selection where redundant features may be removed.
* Data transforms where attributes are scaled or redistributed in order to best expose the structure of the problem later to learning algorithms.

```{r}
email_notify(paste("Data Cleaning and Transformation has begun!",date()))
```

### 3.a) Data Cleaning

```{r}
# Not applicable for this iteration of the project.

# Mark missing values
#invalid <- 0
#entireDataset$some_col[entireDataset$some_col==invalid] <- NA

# Impute missing values
#entireDataset$some_col <- with(entireDataset, impute(some_col, mean))
```

### 3.b) Feature Selection

```{r}
# Not applicable for this iteration of the project.
```

### 3.c) Data Transforms

```{r}
# Not applicable for this iteration of the project.
```

### 3.d) Display the Final Dataset for Model-Building

```{r}
dim(xy_train)
```

```{r}
sapply(xy_train, class)
```

```{r}
email_notify(paste("Data Cleaning and Transformation completed!",date()))
proc.time()-startTimeScript
```

## 4. Model and Evaluate Algorithms
After the data-prep, we next work on finding a workable model by evaluating a subset of machine learning algorithms that are good at exploiting the structure of the training. The typical evaluation tasks include:

* Defining test options such as cross validation and the evaluation metric to use.
* Spot checking a suite of linear and nonlinear machine learning algorithms.
* Comparing the estimated accuracy of algorithms.

For this project, we will evaluate one linear, three non-linear, and three ensemble algorithms:

Linear Algorithm: Linear Discriminant Analysis

Non-Linear Algorithms: Decision Trees (CART), k-Nearest Neighbors, and Support Vector Machine

Ensemble Algorithms: Bagged CART, Random Forest, and Stochastic Gradient Boosting

The random number seed is reset before each run to ensure that the evaluation of each algorithm is performed using the same data splits. It ensures the results are directly comparable.

### 4.a) Generate models using linear algorithms

```{r LDA}
# Linear Discriminant Analysis (Classification)
email_notify(paste("Linear Discriminant Analysis modeling has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
fit.lda <- train(targetVar~., data=xy_train, method="lda", metric=metricTarget, trControl=control)
print(fit.lda)
proc.time()-startTimeModule
email_notify(paste("Linear Discriminant Analysis modeling completed!",date()))
```

### 4.b) Generate models using nonlinear algorithms

```{r CART}
# Decision Tree - CART (Regression/Classification)
email_notify(paste("Decision Tree modeling has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
fit.cart <- train(targetVar~., data=xy_train, method="rpart", metric=metricTarget, trControl=control)
print(fit.cart)
proc.time()-startTimeModule
email_notify(paste("Decision Tree modeling completed!",date()))
```

```{r KNN}
# k-Nearest Neighbors (Regression/Classification)
email_notify(paste("k-Nearest Neighbors modeling has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
fit.knn <- train(targetVar~., data=xy_train, method="knn", metric=metricTarget, trControl=control)
print(fit.knn)
proc.time()-startTimeModule
email_notify(paste("k-Nearest Neighbors modeling completed!",date()))
```

```{r SVM}
# Support Vector Machine (Regression/Classification)
email_notify(paste("Support Vector Machine modeling has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
fit.svm <- train(targetVar~., data=xy_train, method="svmRadial", metric=metricTarget, trControl=control)
print(fit.svm)
proc.time()-startTimeModule
email_notify(paste("Support Vector Machine modeling completed!",date()))
```

### 4.c) Generate models using ensemble algorithms
In this section, we will explore the use and tuning of ensemble algorithms to see whether we can improve the results.

```{r BAGCART}
# Bagged CART (Regression/Classification)
email_notify(paste("Bagged CART modeling has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
fit.bagcart <- train(targetVar~., data=xy_train, method="treebag", metric=metricTarget, trControl=control)
print(fit.bagcart)
proc.time()-startTimeModule
email_notify(paste("Bagged CART modeling completed!",date()))
```

```{r RF}
# Random Forest (Regression/Classification)
email_notify(paste("Random Forest modeling has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
fit.rf <- train(targetVar~., data=xy_train, method="rf", metric=metricTarget, trControl=control)
print(fit.rf)
proc.time()-startTimeModule
email_notify(paste("Random Forest modeling completed!",date()))
```

```{r GBM}
# Stochastic Gradient Boosting (Regression/Classification)
email_notify(paste("Stochastic Gradient Boosting modeling has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
fit.gbm <- train(targetVar~., data=xy_train, method="gbm", metric=metricTarget, trControl=control, verbose=F)
print(fit.gbm)
proc.time()-startTimeModule
email_notify(paste("Stochastic Gradient Boosting modeling completed!",date()))
```

### 4.d) Compare baseline algorithms

```{r SPOT_CHECK}
results <- resamples(list(LDA=fit.lda, CART=fit.cart, kNN=fit.knn, SVM=fit.svm, BagCART=fit.bagcart, RF=fit.rf, GBM=fit.gbm))
summary(results)
dotplot(results)
cat('The average accuracy from all models is:',
    mean(c(results$values$`LDA~Accuracy`,results$values$`CART~Accuracy`,results$values$`kNN~Accuracy`,results$values$`SVM~Accuracy`,results$values$`BagCART~Accuracy`,results$values$`RF~Accuracy`,results$values$`GBM~Accuracy`)))
```

## 5. Improve Accuracy or Results
After we achieve a short list of machine learning algorithms with good level of accuracy, we can leverage ways to improve the accuracy of the models.

Using the three best-perfoming algorithms from the previous section, we will Search for a combination of parameters for each algorithm that yields the best results.

### 5.a) Algorithm Tuning
Finally, we will tune the best-performing algorithms from each group further and see whether we can get more accuracy out of them.

```{r FINAL1}
# Tuning algorithm #1 - Random Forest
email_notify(paste("Algorithm #1 tuning has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
grid <- expand.grid(mtry = c(1:5))
fit.final1 <- train(targetVar~., data=xy_train, method="rf", metric=metricTarget, tuneGrid=grid, trControl=control)
plot(fit.final1)
print(fit.final1)
proc.time()-startTimeModule
email_notify(paste("Algorithm #1 tuning completed!",date()))
```

```{r FINAL2}
# Tuning algorithm #2 - Stochastic Gradient Boosting
email_notify(paste("Algorithm #2 tuning has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
grid <- expand.grid(.n.trees=c(100, 200, 300), .shrinkage=0.1, .interaction.depth=c(1, 2, 3), .n.minobsinnode=10)
fit.final2 <- train(targetVar~., data=xy_train, method="gbm", metric=metricTarget, tuneGrid=grid, trControl=control, verbose=F)
plot(fit.final2)
print(fit.final2)
proc.time()-startTimeModule
email_notify(paste("Algorithm #2 tuning completed!",date()))
```

### 5.d) Compare Algorithms After Tuning

```{r POST_TUNING}
results <- resamples(list(RF=fit.final1, GBM=fit.final2))
summary(results)
dotplot(results)
```

## 6. Finalize Model and Present Results
Once we have narrow down to a model that we believe can make accurate predictions on unseen data, we are ready to finalize it. Finalizing a model may involve sub-tasks such as:

* Using an optimal model tuned to make predictions on unseen data.
* Creating a standalone model using the tuned parameters
* Saving an optimal model to file for later use.

```{r}
email_notify(paste("Model Validation and Final Model Creation has begun!",date()))
```

### 6.a) Predictions on validation dataset

```{r PREDICT}
predictions <- predict(fit.final1, newdata=xy_test)
confusionMatrix(predictions, y_test)
```

### 6.b) Create standalone model on entire training dataset

```{r}
startTimeModule <- proc.time()
library(randomForest)
set.seed(seedNum)

# Combining the training and test datasets to form the original dataset that will be used for training the final model
xy_train <- rbind(xy_train, xy_test)

finalModel <- randomForest(targetVar~., xy_train, mtry=4, na.action=na.omit)
summary(finalModel)
proc.time()-startTimeModule
```

### 6.c) Save model for later use

```{r}
#saveRDS(finalModel, "./finalModel_MultiClass.rds")
```

```{r}
email_notify(paste("Model Validation and Final Model Creation Completed!",date()))
proc.time()-startTimeScript
```
