---
title: "Binary Classification Model for [Project Name] Using R version 11"
author: "David Lowe"
date: "August 2, 2019"
output: 
  html_document: 
    toc: yes
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Template Credit: Adapted from a template made available by Dr. Jason Brownlee of Machine Learning Mastery. [http://machinelearningmastery.com/]

SUMMARY: [Sample Paragraph - The purpose of this project is to construct a predictive model using various machine learning algorithms and to document the end-to-end steps using a template. The Connectionist Bench dataset is a binary classification situation where we are trying to predict one of the two possible outcomes.]

INTRODUCTION: [Sample Paragraph - The data file patterns obtained by bouncing sonar signals off a metal cylinder or a rock at various angles and under various conditions. The transmitted sonar signal is a frequency-modulated chirp, rising in frequency. The data set contains signals obtained from a variety of different aspect angles, spanning 90 degrees for the cylinder and 180 degrees for the rock. Each pattern is a set of 60 numbers in the range 0.0 to 1.0. Each number represents the energy within a particular frequency band, integrated over a certain period of time.]

ANALYSIS: [Sample Paragraph - The baseline performance of the machine learning algorithms achieved an average accuracy of 78.39%. Two algorithms (Random Forest and Gradient Boosting) achieved the top accuracy metrics after the first round of modeling. After a series of tuning trials, Random Forest turned in the top overall result and achieved an accuracy metric of 85.44%. After applying the optimized parameters, the Random Forest algorithm processed the testing dataset with an accuracy of 80.39%, which was below the prediction from the training data.]

CONCLUSION: [Sample Paragraph - For this iteration, the Random Forest algorithm achieved the best overall training and validation results. For this dataset, the Random Forest algorithm could be considered for further modeling.]

Dataset Used: [Connectionist Bench (Sonar, Mines vs. Rocks) Data Set]

Dataset ML Model: Binary classification with [numerical | categorical] attributes

Dataset Reference: [https://archive.ics.uci.edu/ml/datasets/Connectionist+Bench+%28Sonar%2C+Mines+vs.+Rocks%29]

One potential source of performance benchmarks: [https://archive.ics.uci.edu/ml/datasets/Connectionist+Bench+%28Sonar%2C+Mines+vs.+Rocks%29]

The project aims to touch on the following areas:

1. Document a predictive modeling problem end-to-end.
2. Explore data cleaning and transformation options
3. Explore non-ensemble and ensemble algorithms for baseline model performance
4. Explore algorithm tuning techniques for improving model performance

Any predictive modeling machine learning project genrally can be broken down into about six major tasks:

1. Prepare Environment
2. Summarize Data
3. Prepare Data
4. Model and Evaluate Algorithms
5. Improve Accuracy or Results
6. Finalize Model and Present Results

## 1. Prepare Environment

### 1.a) Load libraries and packages

```{r}
startTimeScript <- proc.time()
library(caret)
library(corrplot)
library(DMwR)
library(Hmisc)
library(mailR)
library(ROCR)
library(stringr)
```

### 1.b) Set up the controlling parameters and functions

```{r}
# Create the random seed number for reproducible results
seedNum <- 888

# Set up the notifyStatus flag to stop sending progress emails (setting to TRUE will send status emails!)
notifyStatus <- FALSE
```

```{r}
# Run algorithms using 10-fold cross validation
control <- trainControl(method="repeatedcv", number=10, repeats=1)
metricTarget <- "Accuracy"
```

```{r}
# Set up the email notification function
email_notify <- function(msg=""){
  sender <- Sys.getenv("MAIL_SENDER")
  receiver <- Sys.getenv("MAIL_RECEIVER")
  gateway <- Sys.getenv("SMTP_GATEWAY")
  smtpuser <- Sys.getenv("SMTP_USERNAME")
  password <- Sys.getenv("SMTP_PASSWORD")
  sbj_line <- "Notification from R Binary Classification Script"
  send.mail(
    from = sender,
    to = receiver,
    subject= sbj_line,
    body = msg,
    smtp = list(host.name = gateway, port = 587, user.name = smtpuser, passwd = password, ssl = TRUE),
    authenticate = TRUE,
    send = TRUE)
}
```

```{r}
if (notifyStatus) email_notify(paste("Library and Data Loading has begun!",date()))
```

### 1.c) Load dataset

```{r}
# Slicing up the document path to get the final destination file name
dataset_path <- 'https://archive.ics.uci.edu/ml/machine-learning-databases/undocumented/connectionist-bench/sonar/sonar.all-data'
doc_path_list <- str_split(dataset_path, "/")
dest_file <- doc_path_list[[1]][length(doc_path_list[[1]])]

if (!file.exists(dest_file)) {
  # Download the document from the website
  cat("Downloading", dataset_path, "as", dest_file, "\n")
  download.file(dataset_path, dest_file, mode = "wb")
  cat(dest_file, "downloaded!\n")
#  unzip(dest_file)
#  cat(dest_file, "unpacked!\n")
}

inputFile <- dest_file
colNames <- paste0("attr",1:60)
colNames <- c(colNames, 'targetVar')
Xy_original <- read.csv(inputFile, sep=',', header=FALSE, col.names = colNames)

# Different ways of reading and processing the input dataset. Saving these for future references.
#X_train <- read.fwf("X_train.txt", widths = widthVector, col.names = colNames)
#y_train <- read.csv("y_train.txt", header = FALSE, col.names = c("targetVar"))
#y_train$targetVar <- as.factor(y_train$targetVar)
#Xy_train <- cbind(X_train, y_train)
```

```{r}
# Take a peek at the dataframe after the import
head(Xy_original)
```

```{r}
sapply(Xy_original, class)
```

```{r}
sapply(Xy_original, function(x) sum(is.na(x)))
```

### 1.d) Data Cleaning

```{r}
# Not applicable for this iteration of the project
# Sample code for performing basic data cleaning tasks

# Dropping features
# Xy_original$column_name <- NULL

# Mark missing values
# invalid <- 0
# Xy_original$column_name[Xy_original$column_name==invalid] <- NA

# Impute missing values
# column_median <- median(Xy_original$column_name, na.rm = TRUE)
# Xy_original$column_name[Xy_original$column_name==0] <- column_median
# Xy_original$column_name <- with(Xy_original, impute(column_name, cholumn_median))

# Convert columns from one data type to another
# Xy_original$column_name <- as.integer(Xy_original$column_name)
# Xy_original$column_name <- as.factor(Xy_original$column_name)
```

```{r}
# Take a peek at the dataframe after the cleaning
head(Xy_original)
```

```{r}
sapply(Xy_original, class)
```

```{r}
sapply(Xy_original, function(x) sum(is.na(x)))
```

### 1.e) Splitting Data into Training and Test Sets

```{r}
# Use variable totCol to hold the number of columns in the dataframe
totCol <- ncol(Xy_original)

# Set up variable totAttr for the total number of attribute columns
totAttr <- totCol-1
```

```{r}
# targetCol variable indicates the column location of the target/class variable
# If the first column, set targetCol to 1. If the last column, set targetCol to totCol
# if (targetCol <> 1) and (targetCol <> totCol), be aware when slicing up the dataframes for visualization! 
targetCol <- totCol

# Standardize the class column to the name of targetVar if applicable
#colnames(Xy_original)[targetCol] <- "targetVar"
#Xy_original$targetVar <- relevel(Xy_original$targetVar,"pos")
```

```{r}
# Create various sub-datasets for visualization and cleaning/transformation operations.
set.seed(seedNum)

# Use 75% of the data to train the models and the remaining for testing/validation
training_index <- createDataPartition(Xy_original$targetVar, p=0.75, list=FALSE)
Xy_train <- Xy_original[training_index,]
Xy_test <- Xy_original[-training_index,]

if (targetCol==1) {
  X_train <- Xy_train[,(targetCol+1):totCol]
  y_train <- Xy_train[,targetCol]
  y_test <- Xy_test[,targetCol]
} else {
  X_train <- Xy_train[,1:(totAttr)]
  y_train <- Xy_train[,totCol]
  y_test <- Xy_test[,totCol]
}
```

### 1.f) Set up the parameters for data visualization

```{r}
# Set up the number of row and columns for visualization display. dispRow * dispCol should be >= totAttr
dispCol <- 4
if (totAttr%%dispCol == 0) {
dispRow <- totAttr%/%dispCol
} else {
dispRow <- (totAttr%/%dispCol) + 1
}
cat("Will attempt to create graphics grid (col x row): ", dispCol, ' by ', dispRow)
```

```{r}
if (notifyStatus) email_notify(paste("Library and Data Loading completed!",date()))
```

## 2. Summarize Data
To gain a better understanding of the data that we have on-hand, we will leverage a number of descriptive statistics and data visualization techniques. The plan is to use the results to consider new questions, review assumptions, and validate hypotheses that we can investigate later with specialized models.

```{r}
if (notifyStatus) email_notify(paste("Data Summarization and Visualization has begun!",date()))
```

### 2.a) Descriptive statistics

#### 2.a.i) Peek at the data itself

```{r}
head(Xy_train)
```

#### 2.a.ii) Dimensions of the dataset

```{r}
dim(Xy_train)
```

#### 2.a.iii) Types of the attribute

```{r}
sapply(Xy_train, class)
```

#### 2.a.iv) Statistical summary of the attributes

```{r}
summary(Xy_train)
```

#### 2.a.v) Summarize the levels of the class attribute

```{r}
cbind(freq=table(y_train), percentage=prop.table(table(y_train))*100)
```

### 2.b) Data visualizations

```{r}
# Boxplots for each attribute
# par(mfrow=c(dispRow,dispCol))
for(i in 1:totAttr) {
	boxplot(X_train[,i], main=names(X_train)[i])
}
```

```{r}
# Histograms each attribute
# par(mfrow=c(dispRow,dispCol))
for(i in 1:totAttr) {
	hist(X_train[,i], main=names(X_train)[i])
}
```

```{r}
# Density plot for each attribute
# par(mfrow=c(dispRow,dispCol))
for(i in 1:totAttr) {
	plot(density(X_train[,i]), main=names(X_train)[i])
}
```

```{r}
# Correlation matrix
correlations <- cor(X_train)
corrplot(correlations, method="circle")
```

```{r}
if (notifyStatus) email_notify(paste("Data Summarization and Visualization completed!",date()))
```

## 3. Prepare Data
Some dataset may require additional preparation activities that will best exposes the structure of the problem and the relationships between the input attributes and the output variable. Some data-prep tasks might include:

* Feature selection where redundant features may be removed.
* Data transforms where attributes are scaled or redistributed in order to best expose the structure of the problem later to learning algorithms.

```{r}
if (notifyStatus) email_notify(paste("Data Cleaning and Transformation has begun!",date()))
```

### 3.a) Data Transforms

```{r}
# Not applicable for this iteration of the project
# Sample code for performing SMOTE transformation on the training data

# set.seed(seedNum)
# Xy_train <- SMOTE(targetVar ~., data=Xy_train, perc.over=200, perc.under=300)
# totCol <- ncol(Xy_train)
# y_train <- Xy_train[,totCol]
# cbind(freq=table(y_train), percentage=prop.table(table(y_train))*100)
```

### 3.b) Feature Selection

```{r}
# Not applicable for this iteration of the project
# Sample Code for finding collinear features (Block #1 of 2)

# Using the correlations calculated previously, we try to find attributes that are highly correlated.
# highlyCorrelated <- findCorrelation(correlations, cutoff=0.85)
# print(highlyCorrelated)
# cat('Number of attributes found to be highly correlated:',length(highlyCorrelated))
```

```{r}
# Sample Code for finding collinear features (Block #2 of 2)

# Removing the highly correlated attributes from the training and validation dataframes
# Xy_train <- Xy_train[, -highlyCorrelated]
# Xy_test <- Xy_test[, -highlyCorrelated]
```

```{r}
# Not applicable for this iteration of the project
# Sample code for performing Attribute Importance Ranking (Block #1 of 3)

# startTimeModule <- proc.time()
# set.seed(seedNum)
# library(gbm)
# model_fs <- train(targetVar~., data=Xy_train, method="gbm", preProcess="scale", trControl=control, verbose=F)
# rankedImportance <- varImp(model_fs, scale=FALSE)
# print(rankedImportance)
# plot(rankedImportance)
```

```{r}
# Sample code for performing Attribute Importance Ranking (Block #2 of 3)

# Set the importance threshold and calculate the list of attributes that don't contribute to the importance threshold
# maxThreshold <- 0.99
# rankedAttributes <- rankedImportance$importance
# rankedAttributes <- rankedAttributes[order(-rankedAttributes$Overall),,drop=FALSE]
# totalWeight <- sum(rankedAttributes)
# i <- 1
# accumWeight <- 0
# exit_now <- FALSE
# while ((i <= totAttr) & !exit_now) {
#   accumWeight = accumWeight + rankedAttributes[i,]
#   if ((accumWeight/totalWeight) >= maxThreshold) {
#     exit_now <- TRUE
#   } else {
#     i <- i + 1
#   }
# }
# lowImportance <- rankedAttributes[(i+1):(totAttr),,drop=FALSE]
# lowAttributes <- rownames(lowImportance)
# cat('Number of attributes contributed to the importance threshold:',i,"\n")
# cat('Number of attributes found to be of low importance:',length(lowAttributes))
```

```{r}
# Sample code for performing Attribute Importance Ranking (Block #3 of 3)

# Removing the unselected attributes from the training and validation dataframes
# Xy_train <- Xy_train[, !(names(Xy_train) %in% lowAttributes)]
# Xy_test <- Xy_test[, !(names(Xy_test) %in% lowAttributes)]
```

```{r}
# Not applicable for this iteration of the project
# Sample code for performing Recursive Feature Elimination (Block #1 of 2)

# startTimeModule <- proc.time()
# set.seed(seedNum)
# rfeCTRL <- rfeControl(functions=rfFuncs, method="cv", number=10)
# rfeResults <- rfe(Xy_train[,1:totAttr], Xy_train[,totCol], sizes=c(30:55), rfeControl=rfeCTRL)
# print(rfeResults)
# rfeAttributes <- predictors(rfeResults)
# cat('Number of attributes identified from the RFE algorithm:',length(rfeAttributes))
# print(rfeAttributes)
# plot(rfeResults, type=c("g", "o"))
```

```{r}
# Sample code for performing Recursive Feature Elimination (Block #2 of 3)

# Removing the unselected attributes from the training and validation dataframes
# rfeAttributes <- c(rfeAttributes,"targetVar")
# Xy_train <- Xy_train[, (names(Xy_train) %in% rfeAttributes)]
# Xy_test <- Xy_test[, (names(Xy_test) %in% rfeAttributes)]
```

### 3.c) Display the Final Datasets for Model-Building

```{r}
# We finalize the training and testing datasets for the modeling activities
dim(Xy_train)
dim(Xy_test)
```

```{r}
if (notifyStatus) email_notify(paste("Data Cleaning and Transformation completed!",date()))
```

```{r}
proc.time()-startTimeScript
```

## 4. Model and Evaluate Algorithms
After the data-prep, we next work on finding a workable model by evaluating a subset of machine learning algorithms that are good at exploiting the structure of the training. The typical evaluation tasks include:

* Defining test options such as cross validation and the evaluation metric to use.
* Spot checking a suite of linear and nonlinear machine learning algorithms.
* Comparing the estimated accuracy of algorithms.

For this project, we will evaluate one linear, one non-linear, and three ensemble algorithms:

Linear Algorithm: Logistic Regression

Non-Linear Algorithm: Decision Trees (CART)

Ensemble Algorithms: Bagged CART, Random Forest, and Gradient Boosting

The random number seed is reset before each run to ensure that the evaluation of each algorithm is performed using the same data splits. It ensures the results are directly comparable.

### 4.a) Generate models using linear algorithms

```{r LR}
# Logistic Regression (Classification)
if (notifyStatus) email_notify(paste("Logistic Regression modeling has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
fit.glm <- train(targetVar~., data=Xy_train, method="glm", metric=metricTarget, trControl=control)
print(fit.glm)
proc.time()-startTimeModule
if (notifyStatus) email_notify(paste("Logistic Regression modeling completed!",date()))
```

### 4.b) Generate models using nonlinear algorithms

```{r CART}
# Decision Tree - CART (Regression/Classification)
if (notifyStatus) email_notify(paste("Decision Tree modeling has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
fit.cart <- train(targetVar~., data=Xy_train, method="rpart", metric=metricTarget, trControl=control)
print(fit.cart)
proc.time()-startTimeModule
if (notifyStatus) email_notify(paste("Decision Tree modeling completed!",date()))
```

### 4.c) Generate models using ensemble algorithms
In this section, we will explore the use and tuning of ensemble algorithms to see whether we can improve the results.

```{r BAGCART}
# Bagged CART (Regression/Classification)
if (notifyStatus) email_notify(paste("Bagged CART modeling has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
fit.bagcart <- train(targetVar~., data=Xy_train, method="treebag", metric=metricTarget, trControl=control)
print(fit.bagcart)
proc.time()-startTimeModule
if (notifyStatus) email_notify(paste("Bagged CART modeling completed!",date()))
```

```{r RF}
# Random Forest (Regression/Classification)
if (notifyStatus) email_notify(paste("Random Forest modeling has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
fit.rf <- train(targetVar~., data=Xy_train, method="rf", metric=metricTarget, trControl=control)
print(fit.rf)
proc.time()-startTimeModule
if (notifyStatus) email_notify(paste("Random Forest modeling completed!",date()))
```

```{r GBM}
# Gradient Boosting (Regression/Classification)
if (notifyStatus) email_notify(paste("Gradient Boosting modeling has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
fit.gbm <- train(targetVar~., data=Xy_train, method="xgbTree", metric=metricTarget, trControl=control, verbose=F)
# fit.gbm <- train(targetVar~., data=Xy_train, method="gbm", metric=metricTarget, trControl=control, verbose=F)
print(fit.gbm)
proc.time()-startTimeModule
if (notifyStatus) email_notify(paste("Gradient Boosting modeling completed!",date()))
```

### 4.d) Compare baseline algorithms

```{r SPOT_CHECK}
results <- resamples(list(LR=fit.glm, CART=fit.cart, BagCART=fit.bagcart, RF=fit.rf, GBM=fit.gbm))
summary(results)
dotplot(results)
cat('The average accuracy from all models is:',
    mean(c(results$values$`LR~Accuracy`,results$values$`CART~Accuracy`,results$values$`BagCART~Accuracy`,results$values$`RF~Accuracy`,results$values$`GBM~Accuracy`)))
```

## 5. Improve Accuracy or Results
After we achieve a short list of machine learning algorithms with good level of accuracy, we can leverage ways to improve the accuracy of the models.

Using the three best-perfoming algorithms from the previous section, we will Search for a combination of parameters for each algorithm that yields the best results.

### 5.a) Algorithm Tuning
Finally, we will tune the best-performing algorithms from each group further and see whether we can get more accuracy out of them.

```{r FINAL1}
# Tuning algorithm #1 - Random Forest
if (notifyStatus) email_notify(paste("Algorithm #1 tuning has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
grid <- expand.grid(mtry=c(2,15,30,45,60))
fit.final1 <- train(targetVar~., data=Xy_train, method="rf", metric=metricTarget, trControl=control)
print(fit.final1)
proc.time()-startTimeModule
if (notifyStatus) email_notify(paste("Algorithm #1 tuning completed!",date()))
```

```{r FINAL2}
# Tuning algorithm #2 - Gradient Boosting
if (notifyStatus) email_notify(paste("Algorithm #2 tuning has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
grid <- expand.grid(nrounds=c(100,200,300,400,500), max_depth=3, eta=0.4, gamma=0, colsample_bytree=0.6, min_child_weight=1, subsample=0.5)
fit.final2 <- train(targetVar~., data=Xy_train, method="xgbTree", metric=metricTarget, tuneGrid=grid, trControl=control, verbose=F)
plot(fit.final2)
print(fit.final2)
proc.time()-startTimeModule
if (notifyStatus) email_notify(paste("Algorithm #2 tuning completed!",date()))
```

### 5.d) Compare Algorithms After Tuning

```{r POST_TUNING}
results <- resamples(list(RF=fit.final1, GBM=fit.final2))
summary(results)
dotplot(results)
```

## 6. Finalize Model and Present Results
Once we have narrow down to a model that we believe can make accurate predictions on unseen data, we are ready to finalize it. Finalizing a model may involve sub-tasks such as:

* Using an optimal model tuned to make predictions on unseen data.
* Creating a standalone model using the tuned parameters
* Saving an optimal model to file for later use.

```{r}
if (notifyStatus) email_notify(paste("Model Validation and Final Model Creation has begun!",date()))
```

### 6.a) Predictions on validation dataset

```{r PREDICT1}
predictions <- predict(fit.final1, newdata=Xy_test)
confusionMatrix(predictions, y_test)

pred <- prediction(as.numeric(predictions), as.numeric(y_test))
perf <- performance(pred, measure = "tpr", x.measure = "fpr")
plot(perf, colorize=TRUE)
auc <- performance(pred, measure = "auc")
auc <- auc@y.values[[1]]
cat('Area under the curve is:', auc)
```

```{r PREDICT2}
predictions <- predict(fit.final2, newdata=Xy_test)
confusionMatrix(predictions, y_test)

pred <- prediction(as.numeric(predictions), as.numeric(y_test))
perf <- performance(pred, measure = "tpr", x.measure = "fpr")
plot(perf, colorize=TRUE)
auc <- performance(pred, measure = "auc")
auc <- auc@y.values[[1]]
cat('Area under the curve is:', auc)
```

### 6.b) Create standalone model on entire training dataset

```{r FINALMODEL}
startTimeModule <- proc.time()
set.seed(seedNum)
library(randomForest)

# Combining datasets to form a complete dataset that will be used to train the final model
Xy_complete <- rbind(Xy_train, Xy_test)

finalModel <- randomForest(targetVar~., Xy_complete, mtry=2, na.action=na.omit)
summary(finalModel)
proc.time()-startTimeModule
```

### 6.c) Save model for later use

```{r}
#saveRDS(finalModel, "./finalModel_BinaryClass.rds")
```

```{r}
if (notifyStatus) email_notify(paste("Model Validation and Final Model Creation Completed!",date()))
```

```{r}
proc.time()-startTimeScript
```
