---
title: "Multi-Class Classification Model for Human Activity Recognition with Smartphone Using R Take 2"
author: "David Lowe"
date: "October 24, 2018"
output: 
  html_document: 
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Template Credit: Adapted from a template made available by Dr. Jason Brownlee of Machine Learning Mastery. [http://machinelearningmastery.com/]

Template Credit: Adapted from a template made available by Dr. Jason Brownlee of Machine Learning Mastery. https://machinelearningmastery.com/

SUMMARY: The purpose of this project is to construct a prediction model using various machine learning algorithms and to document the end-to-end steps using a template. The Human Activities with Smartphone Dataset is a multi-class classification situation where we are trying to predict one of the six possible outcomes.

INTRODUCTION: Researchers collected the datasets from experiments that consist of a group of 30 volunteers with each person performed six activities wearing a smartphone on the waist. With its embedded accelerometer and gyroscope, the research captured measurement for the activities of WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING. The dataset has been randomly partitioned into two sets, where 70% of the volunteers were selected for generating the training data and 30% the test data. 

In iteration Take1, the script focuses on evaluating various machine learning algorithms and identify the algorithm that produces the best accuracy metric. Iteration Take1 established a baseline performance regarding accuracy and processing time. For this iteration, we will examine the feasibility of using dimensionality reduction techniques to reduce the processing time while still maintaining an adequate level of prediction accuracy. The first technique we will explore is to eliminate collinear attributes based on a threshold of 85%.

CONCLUSION: From the previous iteration Take 1, the baseline performance of the ten algorithms achieved an average accuracy of 91.67%. Three algorithms (Linear Discriminant Analysis, Support Vector Machine, and Stochastic Gradient Boosting) achieved the top three accuracy scores after the first round of modeling. After a series of tuning trials, Stochastic Gradient Boosting turned in the top result using the training data. It achieved an average accuracy of 98.84%. Stochastic Gradient Boosting also processed the validation dataset with an accuracy of 95.49%, which was slightly below the accuracy from the training data.

From the current iteration, the baseline performance of the ten algorithms achieved an average accuracy of 90.83%. Three algorithms (Random Forest, Support Vector Machine, and Stochastic Gradient Boosting) achieved the top three accuracy scores after the first round of modeling. After a series of tuning trials, Stochastic Gradient Boosting turned in the top result using the training data. It achieved an average accuracy of 98.07%. Using the optimized tuning parameter available, the Stochastic Gradient Boosting algorithm processed the validation dataset with an accuracy of 93.96%, which was slightly worse than the accuracy from the training data and possibly due to over-fitting.

From the model-building activities, the number of attributes went from 561 down to 192 after eliminating 369 variables that are at least 85% collinear. The processing time went from 21 hours 43 minutes in iteration Take1 down to 6 hours and 31 minutes in iteration Take2. That was a reduction in model training and processing time of 70%.

In conclusion, the reduction in the number of attributes used still achieved an acceptable level of accuracy. by reducing the collinearity, the modeling took a much shorter time to process yet still retained decent accuracy. For this dataset, the Stochastic Gradient Boosting algorithm should be considered for further modeling or production use.

Dataset Used: Human Activity Recognition Using Smartphone Data Set

Dataset ML Model: Multi-class classification with numerical attributes

Dataset Reference: https://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones

One potential source of performance benchmarks: https://www.kaggle.com/uciml/human-activity-recognition-with-smartphones

The project aims to touch on the following areas:

1. Document a predictive modeling problem end-to-end.
2. Explore data cleaning and transformation options
3. Explore non-ensemble and ensemble algorithms for baseline model performance
4. Explore algorithm tuning techniques for improving model performance

Any predictive modeling machine learning project genrally can be broken down into about six major tasks:

1. Prepare Problem
2. Summarize Data
3. Prepare Data
4. Model and Evaluate Algorithms
5. Improve Accuracy or Results
6. Finalize Model and Present Results

## 1. Prepare Problem

### 1.a) Load libraries

```{r}
startTimeScript <- proc.time()
library(caret)
library(corrplot)
library(parallel)

# Create one random seed number for reproducible results
seedNum <- 888
set.seed(seedNum)
```

### 1.b) Load dataset

```{r}
widthVector <- rep(16, 561)
colNames <- paste0("attr",1:561)
x_train <- read.fwf("X_train.txt", widths = widthVector, col.names = colNames)
y_train <- read.csv("y_train.txt", header = FALSE, col.names = c("targetVar"))
y_train$targetVar <- as.factor(y_train$targetVar)
xy_train <- cbind(x_train, y_train)

x_test <- read.fwf("X_test.txt", widths = widthVector, col.names = colNames)
y_test <- read.csv("y_test.txt", header = FALSE, col.names = c("targetVar"))
y_test$targetVar <- as.factor(y_test$targetVar)
xy_test <- cbind(x_test, y_test)
```

### 1.c) Set up the key parameters to be used in the script

```{r}
# Use variable totCol to hold the number of columns in the dataframe
totCol <- ncol(xy_train)
```

```{r}
# Set up a variable for the total number of attribute columns (totAttr)
totAttr <- totCol-1
# Set up the number of row and columns for visualization display. dispRow * dispCol should be >= totAttr
dispCol <- 5
if (totAttr%%dispCol == 0) {
dispRow <- totAttr%/%dispCol
} else {
dispRow <- (totAttr%/%dispCol) + 1
}
cat("Will attempt to create graphics grid (col x row): ", dispCol, ' by ', dispRow)
```

```{r}
# Run algorithms using 10-fold cross validation
control <- trainControl(method="repeatedcv", number=10, repeats=3)
metricTarget <- "Accuracy"
```

## 2. Summarize Data
To gain a better understanding of the data that we have on-hand, we will leverage a number of descriptive statistics and data visualization techniques. The plan is to use the results to consider new questions, review assumptions, and validate hypotheses that we can investigate later with specialized models.

### 2.a) Descriptive statistics

#### 2.a.i) Peek at the data itself.

```{r}
head(xy_train)
```

#### 2.a.ii) Dimensions of the dataset.

```{r}
dim(xy_train)
```

#### 2.a.iii) Types of the attributes.

```{r}
sapply(xy_train, class)
```

#### 2.a.iv) Statistical summary of all attributes.

```{r}
summary(xy_train)
```

#### 2.a.v) Summarize the levels of the class attribute.

```{r}
cbind(freq=table(y_train), percentage=prop.table(table(y_train))*100)
```

#### 2.a.vi) Count missing values.

```{r}
sapply(xy_train, function(x) sum(is.na(x)))
```

### 2.b) Data visualizations

#### 2.b.i) Univariate plots to better understand each attribute.

```{r}
# Boxplots for each attribute
#par(mfrow=c(dispRow,dispCol))
#for(i in 1:totAttr) {
#	boxplot(xy_train[,i], main=names(xy_train)[i])
#}
```

```{r}
# Histograms each attribute
#par(mfrow=c(dispRow,dispCol))
#for(i in 1:totAttr) {
#	hist(xy_train[,i], main=names(xy_train)[i])
#}
```

```{r}
# Density plot for each attribute
#par(mfrow=c(dispRow,dispCol))
#for(i in 1:totAttr) {
#	plot(density(xy_train[,i]), main=names(xy_train)[i])
#}
```

#### 2.b.ii) Multivariate plots to better understand the relationships between attributes

```{r}
# Scatterplot matrix colored by class
#pairs(targetVar~., data=xy_train, col=xy_train$targetVar)
```

```{r}
# Box and whisker plots for each attribute by class
#scales <- list(x=list(relation="free"), y=list(relation="free"))
#featurePlot(x=xy_train_x, y=xy_train_y, plot="box", scales=scales)
```

```{r}
# Density plots for each attribute by class value
#featurePlot(x=xy_train_x, y=xy_train_y, plot="density", scales=scales)
```

```{r}
# Correlation plot
correlations <- cor(x_train)
corrplot(correlations, method="circle")
```

## 3. Prepare Data
Some dataset may require additional preparation activities that will best exposes the structure of the problem and the relationships between the input attributes and the output variable. Some data-prep tasks might include:

* Cleaning data by removing duplicates, marking missing values and even imputing missing values.
* Feature selection where redundant features may be removed.
* Data transforms where attributes are scaled or redistributed in order to best expose the structure of the problem later to learning algorithms.

### 3.a) Data Cleaning

```{r}
# Not applicable for this iteration of the project.

# Mark missing values
#invalid <- 0
#xy_train$some_col[xy_train$some_col==invalid] <- NA

# Impute missing values
#xy_train$some_col <- with(xy_train, impute(some_col, mean))
```

### 3.b) Feature Selection

```{r}
# Using the correlations calculated previously, we try to find attributes that are highly correlated.
highlyCorrelated <- findCorrelation(correlations, cutoff=0.85)
print(highlyCorrelated)
cat('Number of attributes found to be highly correlated:',length(highlyCorrelated))
```

```{r}
# Removing the highly correlated attributes from the training and validation dataframes
xy_train <- xy_train[, -highlyCorrelated]
xy_test <- xy_test[, -highlyCorrelated]
```

### 3.c) Data Transforms

```{r}
# Not applicable for this iteration of the project.
```

### 3.d) Split-out training and validation datasets
We create a training dataset (variable name "xy_train") and a validation dataset (variable name "xy_test").

```{r}
set.seed(seedNum)

# Create a list of the rows in the original dataset we can use for training
#training_index <- createDataPartition(xy_train$targetVar, p=0.70, list=FALSE)
# Use 70% of data to training and testing the models
#xy_train <- xy_train[training_index,]
# Select the remaining 30% of the data for validation
#xy_test <- xy_train[-training_index,]

dim(xy_train)
dim(xy_test)
```

```{r}
proc.time()-startTimeScript
```

## 4. Model and Evaluate Algorithms
After the data-prep, we next work on finding a workable model by evaluating a subset of machine learning algorithms that are good at exploiting the structure of the training. The typical evaluation tasks include:

* Defining test options such as cross validation and the evaluation metric to use.
* Spot checking a suite of linear and nonlinear machine learning algorithms.
* Comparing the estimated accuracy of algorithms.

For this project, we will evaluate one linear, three non-linear, and three ensemble algorithms:

Linear Algorithm: Linear Discriminant Analysis

Non-Linear Algorithms: Decision Trees (CART), k-Nearest Neighbors, and Support Vector Machine

Ensemble Algorithms: Bagged CART, Random Forest, and Stochastic Gradient Boosting

The random number seed is reset before each run to ensure that the evaluation of each algorithm is performed using the same data splits. It ensures the results are directly comparable.

### 4.a) Generate models using linear algorithms

```{r LDA}
# Linear Discriminant Analysis (Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.lda <- train(targetVar~., data=xy_train, method="lda", metric=metricTarget, trControl=control)
proc.time()-startTimeModule
```

### 4.b) Generate models using nonlinear algorithms

```{r CART}
# Decision Tree - CART (Regression/Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.cart <- train(targetVar~., data=xy_train, method="rpart", metric=metricTarget, trControl=control)
proc.time()-startTimeModule
```

```{r KNN}
# k-Nearest Neighbors (Regression/Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.knn <- train(targetVar~., data=xy_train, method="knn", metric=metricTarget, trControl=control)
proc.time()-startTimeModule
```

```{r SVM}
# Support Vector Machine (Regression/Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.svm <- train(targetVar~., data=xy_train, method="svmRadial", metric=metricTarget, trControl=control)
proc.time()-startTimeModule
```

### 4.c) Generate models using ensemble algorithms
In this section, we will explore the use and tuning of ensemble algorithms to see whether we can improve the results.

```{r BAGCART}
# Bagged CART (Regression/Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.bagcart <- train(targetVar~., data=xy_train, method="treebag", metric=metricTarget, trControl=control)
proc.time()-startTimeModule
```

```{r RF}
# Random Forest (Regression/Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.rf <- train(targetVar~., data=xy_train, method="rf", metric=metricTarget, trControl=control)
proc.time()-startTimeModule
```

```{r GBM}
# Stochastic Gradient Boosting (Regression/Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.gbm <- train(targetVar~., data=xy_train, method="gbm", metric=metricTarget, trControl=control, verbose=F)
proc.time()-startTimeModule
```

### 4.d) Compare baseline algorithms

```{r SPOT_CHECK}
results <- resamples(list(LDA=fit.lda, CART=fit.cart, kNN=fit.knn, SVM=fit.svm, BagCART=fit.bagcart, RF=fit.rf, GBM=fit.gbm))
summary(results)
dotplot(results)
cat('The average accuracy from all models is:',
    mean(c(results$values$`LR~Accuracy`,results$values$`CART~Accuracy`,results$values$`kNN~Accuracy`,results$values$`SVM~Accuracy`,results$values$`BagCART~Accuracy`,results$values$`RF~Accuracy`,results$values$`GBM~Accuracy`)))
```

## 5. Improve Accuracy or Results
After we achieve a short list of machine learning algorithms with good level of accuracy, we can leverage ways to improve the accuracy of the models.

Using the top two best-perfoming algorithms from the previous section, we will Search for a combination of parameters for each algorithm that yields the best results.

### 5.a) Algorithm Tuning
Finally, we will tune the best-performing algorithms from each group further and see whether we can get more accuracy out of them.

```{r FINAL1}
# Tuning algorithm #1 - Random Forest
startTimeModule <- proc.time()
set.seed(seedNum)
grid <- expand.grid(mtry = c(5:9))
fit.final1 <- train(targetVar~., data=xy_train, method="rf", metric=metricTarget, tuneGrid=grid, trControl=control)
print(fit.final1)
proc.time()-startTimeModule
```

```{r FINAL3}
# Tuning algorithm #2 - Stochastic Gradient Boosting
startTimeModule <- proc.time()
set.seed(seedNum)
grid <- expand.grid(.n.trees=c(800, 900, 1000, 1100, 1200), .shrinkage=c(0.1), .interaction.depth=c(1), .n.minobsinnode=c(10))
fit.final2 <- train(targetVar~., data=xy_train, method="gbm", metric=metricTarget, tuneGrid=grid, trControl=control, verbose=F)
plot(fit.final2)
print(fit.final2)
proc.time()-startTimeModule
```

### 5.d) Compare Algorithms After Tuning

```{r POST_TUNING}
results <- resamples(list(RF=fit.final1, GBM=fit.final2))
summary(results)
dotplot(results)
```

## 6. Finalize Model and Present Results
Once we have narrow down to a model that we believe can make accurate predictions on unseen data, we are ready to finalize it. Finalizing a model may involve sub-tasks such as:

* Using an optimal model tuned to make predictions on unseen data.
* Creating a standalone model using the tuned parameters
* Saving an optimal model to file for later use.

### 6.a) Predictions on validation dataset

```{r PREDICT}
predictions <- predict(fit.final2, newdata=xy_test)
confusionMatrix(predictions, xy_test$targetVar)
```

### 6.b) Create standalone model on entire training dataset

```{r}
startTimeModule <- proc.time()
library(gbm)
set.seed(seedNum)
finalModel <- gbm(targetVar~., data=xy_train, n.trees=1200, verbose=F)
summary(finalModel)
proc.time()-startTimeModule
```

### 6.c) Save model for later use

```{r}
#saveRDS(finalModel, "./finalModel_MultiClass.rds")
```

```{r}
proc.time()-startTimeScript
```
