---
title: "Binary Classification Model for Census Income Using R Take 2"
author: "David Lowe"
date: "August 8, 2018"
output: 
  html_document: 
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Template Credit: Adapted from a template made available by Dr. Jason Brownlee of Machine Learning Mastery. [http://machinelearningmastery.com/]

SUMMARY: The purpose of this project is to construct a prediction model using various machine learning algorithms and to document the end-to-end steps using a template. The Census Income dataset is a classic binary classification situation where we are trying to predict one of the two possible outcomes.

INTRODUCTION: This data was extracted from the 1994 Census Bureau database by Ronny Kohavi and Barry Becker (Data Mining and Visualization, Silicon Graphics). A set of reasonably clean records was extracted using the following conditions: ((AAGE>16) && (AGI>100) && (AFNLWGT>1) && (HRSWK>0)). The prediction task is to determine whether a person makes over 50K a year.

This dataset has many cells with missing values, so we will examine the models by imputing the missing cells with a default value. This iteration of the project will produce a set of results that we will use to compare with the baseline models from Take 1.

CONCLUSION: From the previous iteration (Take 1), the baseline performance of the ten algorithms achieved an average accuracy of 83.79%. Three ensemble algorithms (Bagged CART, Random Forest, and Stochastic Gradient Boosting) achieved the top accuracy scores after the first round of modeling. After a series of tuning trials, Stochastic Gradient Boosting turned in the top result using the training data. It achieved an average accuracy of 86.27%. Using the optimized tuning parameter available, the Stochastic Gradient Boosting algorithm further processed the validation dataset with an accuracy of 86.29%, which was on-par with the accuracy of the training data.

From this iteration (Take 2), the baseline performance of the ten algorithms achieved an average accuracy of 84.19%. Three ensemble algorithms (Bagged CART, Random Forest, and Stochastic Gradient Boosting) achieved the top accuracy scores after the first round of modeling. After a series of tuning trials, Stochastic Gradient Boosting turned in the top result using the training data. It achieved an average accuracy of 86.60%. Using the optimized tuning parameter available, the Stochastic Gradient Boosting algorithm further processed the validation dataset with an accuracy of 86.93%, which was slightly better than the accuracy of the training data.

For this project, imputing the missing values appeared to have contributed to a slight improvement of the overall accuracy of the training model. The Stochastic Gradient Boosting ensemble algorithm continued to yield consistently top-notch training and validation results, which warrant the additional processing required by the algorithm.

Dataset Used: Census Income Data Set

Dataset ML Model: Binary classification with numerical and categorical attributes

Dataset Reference: https://archive.ics.uci.edu/ml/datasets/Census+Income

One potential source of performance benchmark: https://www.kaggle.com/uciml/adult-census-income

The project aims to touch on the following areas:

1. Document a predictive modeling problem end-to-end.
2. Explore data cleaning and transformation options
3. Explore non-ensemble and ensemble algorithms for baseline model performance
4. Explore algorithm tuning techniques for improving model performance

Any predictive modeling machine learning project genrally can be broken down into about six major tasks:

1. Prepare Problem
2. Summarize Data
3. Prepare Data
4. Model and Evaluate Algorithms
5. Improve Accuracy or Results
6. Finalize Model and Present Results

## 1. Prepare Problem

### 1.a) Load libraries

```{r}
library(caret)
library(corrplot)
library(ROCR)
```

### 1.b) Load dataset

```{r}
startTimeScript <- proc.time()

# Import the training dataset, supply the attribue names, and strip the blanks from the cells
entireDataset <- read.csv("adult.data", sep=",", header=FALSE, na.strings=" ?", stringsAsFactors=FALSE)
colnames(entireDataset) <- c("age", "workclass", "fnlwgt", "education", "education_num", "marital_status", "occupation", "relationship", "race", "sex", "capital_gain", "capital_loss", "hours_per_week", "native_country", "income")
entireDataset$workclass <- gsub('\\s+', '', entireDataset$workclass)
entireDataset$education <- gsub('\\s+', '', entireDataset$education)
entireDataset$marital_status <- gsub('\\s+', '', entireDataset$marital_status)
entireDataset$occupation <- gsub('\\s+', '', entireDataset$occupation)
entireDataset$relationship <- gsub('\\s+', '', entireDataset$relationship)
entireDataset$race <- gsub('\\s+', '', entireDataset$race)
entireDataset$sex <- gsub('\\s+', '', entireDataset$sex)
entireDataset$native_country <- gsub('\\s+', '', entireDataset$native_country)
entireDataset$income <- gsub('\\s+', '', entireDataset$income)

# Import the test dataset, supply the attribue names, and strip the blanks from the cells
testDataset <- read.csv("adult.test", sep=",", header=TRUE, na.strings=" ?", stringsAsFactors=FALSE)
colnames(testDataset) <- c("age", "workclass", "fnlwgt", "education", "education_num", "marital_status", "occupation", "relationship", "race", "sex", "capital_gain", "capital_loss", "hours_per_week", "native_country", "income")
testDataset$workclass <- gsub('\\s+', '', testDataset$workclass)
testDataset$education <- gsub('\\s+', '', testDataset$education)
testDataset$marital_status <- gsub('\\s+', '', testDataset$marital_status)
testDataset$occupation <- gsub('\\s+', '', testDataset$occupation)
testDataset$relationship <- gsub('\\s+', '', testDataset$relationship)
testDataset$race <- gsub('\\s+', '', testDataset$race)
testDataset$sex <- gsub('\\s+', '', testDataset$sex)
testDataset$native_country <- gsub('\\s+', '', testDataset$native_country)
testDataset$income <- gsub('\\s+', '', testDataset$income)

# Rename the class/target column to a standard label
colnames(entireDataset)[ncol(entireDataset)] <- "targetVar"
colnames(testDataset)[ncol(testDataset)] <- "targetVar"

# Use variable totCol to hold the number of columns in the dataframe
totCol <- ncol(entireDataset)
```

## 2. Summarize Data
To gain a better understanding of the data that we have on-hand, we will leverage a number of descriptive statistics and data visualization techniques. The plan is to use the results to consider new questions, review assumptions, and validate hypotheses that we can investigate later with specialized models.

### 2.a) Descriptive statistics

#### 2.a.i) Peek at the data itself.

```{r}
head(entireDataset)
```

#### 2.a.ii) Dimensions of the dataset.

```{r}
dim(entireDataset)
```

#### 2.a.iii) Types of the attributes.

```{r}
sapply(entireDataset, class)
```

#### 2.a.iv) Statistical summary of all attributes.

```{r}
summary(entireDataset)
```

#### 2.a.v) Summarize the levels of the class attribute.

```{r}
entireDataset_x <- entireDataset[,1:(totCol-1)]
entireDataset_y <- entireDataset[,totCol]
cbind(freq=table(entireDataset_y), percentage=prop.table(table(entireDataset_y))*100)
```

#### 2.a.vi) Count missing values.

```{r}
sapply(entireDataset, function(x) sum(is.na(x)))
```

### 2.b) Data visualizations

#### 2.b.i) Univariate plots to better understand each attribute.

```{r}
# Set up a variable for the total number of attribute columns (totAttr)
totAttr <- totCol-1
# Set up the number of row and columns for visualization display. dispRow * dispCol should be >= totAttr
dispCol <- 4
if (totAttr%%dispCol == 0) {
dispRow <- totAttr%/%dispCol
} else {
dispRow <- (totAttr%/%dispCol) + 1
}
cat("Will attempt to create graphics grid (col x row): ", dispCol, ' by ', dispRow)
```


```{r}
# Boxplots for each attribute
#par(mfrow=c(dispRow,dispCol))
#for(i in 1:totAttr) {
#	boxplot(entireDataset[,i], main=names(entireDataset)[i])
#}
```

```{r}
# Histograms each attribute
#par(mfrow=c(dispRow,dispCol))
#for(i in 1:totAttr) {
#	hist(entireDataset[,i], main=names(entireDataset)[i])
#}
```

```{r}
# Density plot for each attribute
#par(mfrow=c(dispRow,dispCol))
#for(i in 1:totAttr) {
#	plot(density(entireDataset[,i]), main=names(entireDataset)[i])
#}
```

#### 2.b.ii) Multivariate plots to better understand the relationships between attributes

```{r}
# Scatterplot matrix colored by class
#pairs(targetVar~., data=entireDataset, col=entireDataset$targetVar)
```

```{r}
# Box and whisker plots for each attribute by class
#scales <- list(x=list(relation="free"), y=list(relation="free"))
#featurePlot(x=entireDataset_x, y=entireDataset_y, plot="box", scales=scales)
```

```{r}
# Density plots for each attribute by class value
#featurePlot(x=entireDataset_x, y=entireDataset_y, plot="density", scales=scales)
```

```{r}
# Correlation plot
#correlations <- cor(entireDataset_x)
#corrplot(correlations, method="circle")
```

## 3. Prepare Data
Some dataset may require additional preparation activities that will best exposes the structure of the problem and the relationships between the input attributes and the output variable. Some data-prep tasks might include:

* Cleaning data by removing duplicates, marking missing values and even imputing missing values.
* Feature selection where redundant features may be removed.
* Data transforms where attributes are scaled or redistributed in order to best expose the structure of the problem later to learning algorithms.

### 3.a) Data Cleaning

```{r}
# First we impute the rows that have missing values
entireDataset$workclass[is.na(entireDataset$workclass)] <- "Not-disclosed"
entireDataset$occupation[is.na(entireDataset$occupation)] <- "Not-disclosed"
entireDataset$native_country[is.na(entireDataset$native_country)] <- "Not-disclosed"
dim(entireDataset)
sapply(entireDataset, function(x) sum(is.na(x)))
```

```{r}
testDataset$workclass[is.na(testDataset$workclass)] <- "Not-disclosed"
testDataset$occupation[is.na(testDataset$occupation)] <- "Not-disclosed"
testDataset$native_country[is.na(testDataset$native_country)] <- "Not-disclosed"
dim(testDataset)
sapply(testDataset, function(x) sum(is.na(x)))
```

```{r}
# As it turned out, the training and test datasets had slightly different class labels.
# The training dataset had <=50K and >50K, where the test dataset had <=50K. and >50K. (with extra period).
# We need to make sure both datasets labels the classes the same way.
testDataset$targetVar[testDataset$targetVar=="<=50K."] <- "<=50K"
testDataset$targetVar[testDataset$targetVar==">50K."] <- ">50K"
```

### 3.b) Feature Selection

```{r}
# Not applicable for this iteration of the project.
```

### 3.c) Data Transforms

```{r}
# Setting some of the columns from string to categorical attributes
entireDataset$workclass <- as.factor(entireDataset$workclass)
entireDataset$education <- as.factor(entireDataset$education)
entireDataset$marital_status <- as.factor(entireDataset$marital_status)
entireDataset$occupation <- as.factor(entireDataset$occupation)
entireDataset$relationship <- as.factor(entireDataset$relationship)
entireDataset$race <- as.factor(entireDataset$race)
entireDataset$sex <- as.factor(entireDataset$sex)
entireDataset$native_country <- as.factor(entireDataset$native_country)
entireDataset$targetVar <- as.factor(entireDataset$targetVar)

testDataset$workclass <- as.factor(testDataset$workclass)
testDataset$education <- as.factor(testDataset$education)
testDataset$marital_status <- as.factor(testDataset$marital_status)
testDataset$occupation <- as.factor(testDataset$occupation)
testDataset$relationship <- as.factor(testDataset$relationship)
testDataset$race <- as.factor(testDataset$race)
testDataset$sex <- as.factor(testDataset$sex)
testDataset$native_country <- as.factor(testDataset$native_country)
testDataset$targetVar <- as.factor(testDataset$targetVar)
```

### 3.d) Split-out training and validation datasets
We create a training dataset (variable name "training") and a validation dataset (variable name "validation").

```{r}
seedNum <- 888
set.seed(seedNum)

# Create a list of the rows in the original dataset we can use for training
#training_index <- createDataPartition(entireDataset$targetVar, p=0.70, list=FALSE)

# Use 70% of data to training and testing the models
training <- entireDataset

# Select the remaining 30% of the data for validation
validation <- testDataset
```

```{r}
proc.time()-startTimeScript
```

## 4. Model and Evaluate Algorithms
After the data-prep, we next work on finding a workable model by evaluating a subset of machine learning algorithms that are good at exploiting the structure of the training. The typical evaluation tasks include:

* Defining test options such as cross validation and the evaluation metric to use.
* Spot checking a suite of linear and nonlinear machine learning algorithms.
* Comparing the estimated accuracy of algorithms.

For this project, we will evaluate one linear, three non-linear, and four ensemble algorithms:

Linear Algorithm: Logistic Regression

Non-Linear Algorithms: Decision Trees (CART), k-Nearest Neighbors, and Support Vector Machine

Ensemble Algorithms: Bagged CART, Random Forest, and Stochastic Gradient Boosting

The random number seed is reset before each run to ensure that the evaluation of each algorithm is performed using the same data splits. It ensures the results are directly comparable.

### 4.a) Set test options and evaluation metric

```{r}
# Run algorithms using 10-fold cross validation
control <- trainControl(method="repeatedcv", number=10, repeats=3)
metricTarget <- "Accuracy"
```

### 4.b) Generate models using linear algorithms

```{r LR}
# Logistic Regression (Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.glm <- train(targetVar~., data=training, method="glm", metric=metricTarget, trControl=control)
proc.time()-startTimeModule
```

### 4.c) Generate models using nonlinear algorithms

```{r CART}
# Decision Tree - CART (Regression/Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.cart <- train(targetVar~., data=training, method="rpart", metric=metricTarget, trControl=control)
proc.time()-startTimeModule
```

```{r KNN}
# k-Nearest Neighbors (Regression/Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.knn <- train(targetVar~., data=training, method="knn", metric=metricTarget, trControl=control)
proc.time()-startTimeModule
```

```{r SVM}
# Support Vector Machine (Regression/Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.svm <- train(targetVar~., data=training, method="svmRadial", metric=metricTarget, trControl=control)
proc.time()-startTimeModule
```

### 4.d) Generate models using ensemble algorithms
In this section, we will explore the use and tuning of ensemble algorithms to see whether we can improve the results.

```{r BAGCART}
# Bagged CART (Regression/Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.bagcart <- train(targetVar~., data=training, method="treebag", metric=metricTarget, trControl=control)
proc.time()-startTimeModule
```

```{r RF}
# Random Forest (Regression/Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.rf <- train(targetVar~., data=training, method="rf", metric=metricTarget, trControl=control)
proc.time()-startTimeModule
```

```{r GBM}
# Stochastic Gradient Boosting (Regression/Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.gbm <- train(targetVar~., data=training, method="gbm", metric=metricTarget, trControl=control, verbose=F)
proc.time()-startTimeModule
```

### 4.e) Compare baseline algorithms

```{r SPOT_CHECK}
results <- resamples(list(LR=fit.glm, CART=fit.cart, kNN=fit.knn, SVM=fit.svm, BagCART=fit.bagcart, RF=fit.rf, GBM=fit.gbm))
summary(results)
dotplot(results)
```

## 5. Improve Accuracy or Results
After we achieve a short list of machine learning algorithms with good level of accuracy, we can leverage ways to improve the accuracy of the models.

Using the three best-perfoming algorithms from the previous section, we will Search for a combination of parameters for each algorithm that yields the best results.

### 5.a) Algorithm Tuning
Finally, we will tune the best-performing algorithms from each group further and see whether we can get more accuracy out of them.

```{r FINAL1}
# Tuning algorithm #1 - Bagged CART
# Bagged CART does not use specialized tuning parameters
startTimeModule <- proc.time()
set.seed(seedNum)
fit.final1 <- fit.bagcart
print(fit.final1)
proc.time()-startTimeModule
```

```{r FINAL2}
# Tuning algorithm #2 - Random Forest
startTimeModule <- proc.time()
set.seed(seedNum)
grid <- expand.grid(mtry = c(2:6))
fit.final2 <- train(targetVar~., data=training, method="rf", metric=metricTarget, tuneGrid=grid, trControl=control)
plot(fit.final2)
print(fit.final2)
proc.time()-startTimeModule
```

```{r FINAL3}
# Tuning algorithm #3 - Stochastic Gradient Boosting
startTimeModule <- proc.time()
set.seed(seedNum)
grid <- expand.grid(.n.trees=c(50, 100, 200), .shrinkage=c(0.01, 0.05, 0.1), .interaction.depth=c(1, 2, 3), .n.minobsinnode=c(3, 5, 7))
fit.final3 <- train(targetVar~., data=training, method="gbm", metric=metricTarget, tuneGrid=grid, trControl=control, verbose=F)
plot(fit.final3)
print(fit.final3)
proc.time()-startTimeModule
```

### 5.d) Compare Algorithms After Tuning

```{r POST_TUNING}
results <- resamples(list(BagCART=fit.final1, RF=fit.final2, GBM=fit.final3))
summary(results)
dotplot(results)
```

## 6. Finalize Model and Present Results
Once we have narrow down to a model that we believe can make accurate predictions on unseen data, we are ready to finalize it. Finalizing a model may involve sub-tasks such as:

* Using an optimal model tuned to make predictions on unseen data.
* Creating a standalone model using the tuned parameters
* Saving an optimal model to file for later use.

### 6.a) Predictions on validation dataset

```{r PREDICT}
predictions <- predict(fit.final3, newdata=validation)
confusionMatrix(predictions, validation$targetVar)

pred <- prediction(as.numeric(predictions), as.numeric(validation$targetVar))
perf <- performance(pred, measure = "tpr", x.measure = "fpr")
plot(perf, colorize=TRUE)
auc <- performance(pred, measure = "auc")
auc <- auc@y.values[[1]]
auc
```

### 6.b) Create standalone model on entire training dataset

```{r FINALMODEL}
startTimeModule <- proc.time()
set.seed(seedNum)
grid <- expand.grid(.n.trees=200, .shrinkage=0.1, .interaction.depth=3, .n.minobsinnode=3)
finalModel <- train(targetVar~., data=entireDataset, method="gbm", metric=metricTarget, tuneGrid=grid, trControl=control, verbose=F)
print(finalModel)
proc.time()-startTimeModule
```

### 6.c) Save model for later use

```{r}
#saveRDS(finalModel, "./finalModel_BinaryClass.rds")
```

```{r}
proc.time()-startTimeScript
```
