---
title: "Binary Classification Model for MiniBooNE Particle Identification Using R Take 4"
author: "David Lowe"
date: "January 21, 2019"
output: 
  html_document: 
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Template Credit: Adapted from a template made available by Dr. Jason Brownlee of Machine Learning Mastery. [https://machinelearningmastery.com/]

SUMMARY: The purpose of this project is to construct a prediction model using various machine learning algorithms and to document the end-to-end steps using a template. The MiniBooNE Particle Identification dataset is a classic binary classification situation where we are trying to predict one of the two possible outcomes.

INTRODUCTION: This dataset is taken from the MiniBooNE experiment and is used to distinguish electron neutrinos (signal) from muon neutrinos (background). The data file is set up as follows. In the first line is the number of signal events followed by the number of background events. The records with the signal events come first, followed by the background events. Each line, after the first line, has the 50 particle ID variables for one event.

From the previous iteration Take1, the script focused on evaluating various machine learning algorithms and identifying the model that produces the best overall metrics. Iteration Take1 established the baseline performance for accuracy and processing time.

From the previous iteration Take2, we examined the feature selection technique of eliminating collinear features. By eliminating the collinear features, we hoped to decrease the processing time and maintain a similar level of accuracy compared to iteration Take1.

From the previous iteration Take3, we examined the feature selection technique of attribute importance ranking. By taking only the most important attributes, we hoped to decrease the processing time and maintain a similar level of accuracy compared to iterations Take1 and Take2.

In the current iteration Take4, we will explore the Recursive Feature Elimination (or RFE) technique by recursively removing attributes and building a model on those attributes that remain.

ANALYSIS: From the previous iteration Take1, the baseline performance of the eight algorithms achieved an average accuracy of 90.82%. Two algorithms (Bagged CART and Random Forest) achieved the top accuracy scores after the first round of modeling. After a series of tuning trials, Random Forest turned in the top result using the training data. It achieved an average accuracy of 93.74%. By optimizing the tuning parameters, the Random Forest algorithm processed the testing dataset with an accuracy of 93.91%, which was even better than the training data.

From the previous iteration Take2, the baseline performance of the machine learning algorithms achieved an average accuracy of 90.04%. Two algorithms (Stochastic Gradient Boosting and Random Forest) achieved the top accuracy metrics after the first round of modeling. After a series of tuning trials, Stochastic Gradient Boosting turned in the top overall result and achieved an accuracy metric of 93.47%. By using the optimized parameters, the Stochastic Gradient Boosting algorithm processed the testing dataset with an accuracy of 93.57%, which was even better than the training data.

From the previous iteration Take3, the baseline performance of the machine learning algorithms achieved an average accuracy of 90.49%. Two algorithms (Stochastic Gradient Boosting and Random Forest) achieved the top accuracy metrics after the first round of modeling. After a series of tuning trials, Random Forest turned in the top overall result and achieved an accuracy metric of 93.52%. By using the optimized parameters, the Stochastic Gradient Boosting algorithm processed the testing dataset with an accuracy of 93.74%, which was even better than the training data.

In the current iteration Take4, the baseline performance of the machine learning algorithms achieved an average accuracy of 90.79%. Two algorithms (Stochastic Gradient Boosting and Random Forest) achieved the top accuracy metrics after the first round of modeling. After a series of tuning trials, Random Forest turned in the top overall result and achieved an accuracy metric of 93.75%. By using the optimized parameters, the Stochastic Gradient Boosting algorithm processed the testing dataset with an accuracy of 93.91%, which was even better than the training data.

From the model-building perspective, the number of attributes decreased by 10, from 50 down to 40 in iteration Take4. The processing time went from 17 hours 18 minutes in iteration Take1 down to 30 hours 58 minutes in Take3, which was an increase of 21.5% from Take1. It was a significant increase in comparison to Take2, which had a processing time of 12 hours 17 minutes. It was also a significant increase in comparison to Take3, which had a processing time of 6 hours 48 minutes.

CONCLUSION: For this iteration, the Random Forest algorithm achieved the best overall results with an increased processing time after running Recursive Feature Elimination. For this dataset, the Stochastic Gradient Boosting and Random Forest algorithms should be considered for further modeling or production use.

Dataset Used: MiniBooNE particle identification Data Set

Dataset ML Model: Binary classification with numerical attributes

Dataset Reference: https://archive.ics.uci.edu/ml/datasets/MiniBooNE+particle+identification

The project aims to touch on the following areas:

1. Document a predictive modeling problem end-to-end.
2. Explore data cleaning and transformation options
3. Explore non-ensemble and ensemble algorithms for baseline model performance
4. Explore algorithm tuning techniques for improving model performance

Any predictive modeling machine learning project genrally can be broken down into about six major tasks:

1. Prepare Problem
2. Summarize Data
3. Prepare Data
4. Model and Evaluate Algorithms
5. Improve Accuracy or Results
6. Finalize Model and Present Results

## 1. Prepare Problem

### 1.a) Load libraries

```{r}
startTimeScript <- proc.time()
library(caret)
library(corrplot)
library(mailR)
library(stringr)
library(MLmetrics)
library(DMwR)
library(ROCR)

# Create one random seed number for reproducible results
seedNum <- 888
set.seed(seedNum)
```

### 1.b) Set up the email notification function

```{r}
email_notify <- function(msg=""){
  sender <- "luozhi2488@gmail.com"
  receiver <- "dave@contactdavidlowe.com"
  sbj_line <- "Notification from R Script"
  password <- readLines("../email_credential.txt")
  send.mail(
    from = sender,
    to = receiver,
    subject= sbj_line,
    body = msg,
    smtp = list(host.name = "smtp.gmail.com", port = 465, user.name = sender, passwd = password, ssl = TRUE),
    authenticate = TRUE,
    send = TRUE)
}
```

```{r}
email_notify(paste("Library and Data Loading has begun!",date()))
```

### 1.c) Load dataset

```{r}
# Process the first line of the input file with the meta information
con <- file("MiniBooNE_PID.txt","r")
first_line <- readLines(con,n=1)
str_split(first_line, " ")[1]
close(con)
signal_rec <- as.numeric(str_split(first_line," ")[[1]][2])
background_rec <- as.numeric(str_split(first_line," ")[[1]][3])

# Process the remainder of the input file and capture the fixed length records
widthVector <- rep(14, 50)
colNames <- paste0("attr",1:50)
xy_original <- read.fwf("MiniBooNE_PID.txt", widths=widthVector, skip=1, col.names=colNames)
xy_original$targetVar <- "background"
xy_original[1:signal_rec,"targetVar"] <- "signal"
xy_original$targetVar <- as.factor(xy_original$targetVar)
xy_original$targetVar <- relevel(xy_original$targetVar,"signal")
```

```{r}
# Use variable totCol to hold the number of columns in the dataframe
totCol <- ncol(xy_original)

# Set up variable totAttr for the total number of attribute columns
totAttr <- totCol-1
```

```{r}
# targetCol variable indicates the column location of the target/class variable
# If the first column, set targetCol to 1. If the last column, set targetCol to totCol
# if (targetCol <> 1) and (targetCol <> totCol), be aware when slicing up the dataframes for visualization! 
targetCol <- totCol
colnames(xy_original)[targetCol] <- "targetVar"
```

```{r}
# We create training datasets (xy_train, x_train, y_train) for various operations.
# We create test datasets (xy_test, x_test, y_test) for various operations.
set.seed(seedNum)

# Create a list of the rows in the original dataset we can use for training
training_index <- createDataPartition(xy_original$targetVar, p=0.70, list=FALSE)
# Use 70% of the data to train the models and the remaining for testing/validation
xy_train <- xy_original[training_index,]
xy_test <- xy_original[-training_index,]

if (targetCol==1) {
x_train <- xy_train[,(targetCol+1):totCol]
y_train <- xy_train[,targetCol]
y_test <- xy_test[,targetCol]
} else {
x_train <- xy_train[,1:(totAttr)]
y_train <- xy_train[,totCol]
y_test <- xy_test[,totCol]
}
```

### 1.d) Set up the key parameters to be used in the script

```{r}
# Set up the number of row and columns for visualization display. dispRow * dispCol should be >= totAttr
dispCol <- 4
if (totAttr%%dispCol == 0) {
dispRow <- totAttr%/%dispCol
} else {
dispRow <- (totAttr%/%dispCol) + 1
}
cat("Will attempt to create graphics grid (col x row): ", dispCol, ' by ', dispRow)
```

### 1.e) Set test options and evaluation metric

```{r}
# Run algorithms using 10-fold cross validation
control <- trainControl(method="repeatedcv", number=10, repeats=1)
metricTarget <- "Accuracy"
```

```{r}
email_notify(paste("Library and Data Loading completed!",date()))
```

## 2. Summarize Data
To gain a better understanding of the data that we have on-hand, we will leverage a number of descriptive statistics and data visualization techniques. The plan is to use the results to consider new questions, review assumptions, and validate hypotheses that we can investigate later with specialized models.

```{r}
email_notify(paste("Data Summarization and Visualization has begun!",date()))
```

### 2.a) Descriptive statistics

#### 2.a.i) Peek at the data itself.

```{r}
head(xy_train)
```

#### 2.a.ii) Dimensions of the dataset.

```{r}
dim(xy_train)
```

#### 2.a.iii) Types of the attributes.

```{r}
sapply(xy_train, class)
```

#### 2.a.iv) Statistical summary of all attributes.

```{r}
summary(xy_train)
```

#### 2.a.v) Summarize the levels of the class attribute.

```{r}
cbind(freq=table(y_train), percentage=prop.table(table(y_train))*100)
```

#### 2.a.vi) Count missing values.

```{r}
sapply(xy_train, function(x) sum(is.na(x)))
```

### 2.b) Data visualizations

#### 2.b.i) Univariate plots to better understand each attribute.

```{r}
# Boxplots for each attribute
# par(mfrow=c(dispRow,dispCol))
for(i in 1:totAttr) {
	boxplot(x_train[,i], main=names(x_train)[i])
}
```

```{r}
# Histograms each attribute
# par(mfrow=c(dispRow,dispCol))
for(i in 1:totAttr) {
	hist(x_train[,i], main=names(x_train)[i])
}
```

```{r}
# Density plot for each attribute
# par(mfrow=c(dispRow,dispCol))
for(i in 1:totAttr) {
	plot(density(x_train[,i]), main=names(x_train)[i])
}
```

#### 2.b.ii) Multivariate plots to better understand the relationships between attributes

```{r}
# Scatterplot matrix colored by class
# pairs(targetVar~., data=xy_train, col=xy_train$targetVar)
```

```{r}
# Box and whisker plots for each attribute by class
scales <- list(x=list(relation="free"), y=list(relation="free"))
featurePlot(x=x_train, y=y_train, plot="box", scales=scales)
```

```{r}
# Density plots for each attribute by class value
featurePlot(x=x_train, y=y_train, plot="density", scales=scales)
```

```{r}
# Correlation plot
correlations <- cor(x_train, method = "spearman")
corrplot(correlations, method="circle")
```

```{r}
email_notify(paste("Data Summarization and Visualization completed!",date()))
```

## 3. Prepare Data
Some dataset may require additional preparation activities that will best exposes the structure of the problem and the relationships between the input attributes and the output variable. Some data-prep tasks might include:

* Cleaning data by removing duplicates, marking missing values and even imputing missing values.
* Feature selection where redundant features may be removed.
* Data transforms where attributes are scaled or redistributed in order to best expose the structure of the problem later to learning algorithms.

```{r}
email_notify(paste("Data Cleaning and Transformation has begun!",date()))
```

### 3.a) Data Cleaning

```{r}
# Not applicable for this iteration of the project.
```

### 3.b) Feature Selection

```{r}
# Sample code for perfoming feature selection using the Recursive Feature Elimination (RFE) technique
startTimeModule <- proc.time()
set.seed(seedNum)
rfeCTRL <- rfeControl(functions=rfFuncs, method="cv", number=10)
rfeResults <- rfe(xy_train[,1:totAttr], xy_train[,totCol], sizes=c(30:55), rfeControl=rfeCTRL)
print(rfeResults)
rfeAttributes <- predictors(rfeResults)
cat('Number of attributes identified from the RFE algorithm:',length(rfeAttributes))
print(rfeAttributes)
plot(rfeResults, type=c("g", "o"))
```

```{r}
# Removing the unselected attributes from the training and validation dataframes
rfeAttributes <- c(rfeAttributes,"targetVar")
xy_train <- xy_train[, (names(xy_train) %in% rfeAttributes)]
xy_test <- xy_test[, (names(xy_test) %in% rfeAttributes)]
```

### 3.c) Data Transforms

```{r}
# Not applicable for this iteration of the project.
```

### 3.d) Display the Final Dataset for Model-Building

```{r}
dim(xy_train)
```

```{r}
sapply(xy_train, class)
```

```{r}
email_notify(paste("Data Cleaning and Transformation completed!",date()))
proc.time()-startTimeScript
```

## 4. Model and Evaluate Algorithms
After the data-prep, we next work on finding a workable model by evaluating a subset of machine learning algorithms that are good at exploiting the structure of the training. The typical evaluation tasks include:

* Defining test options such as cross validation and the evaluation metric to use.
* Spot checking a suite of linear and nonlinear machine learning algorithms.
* Comparing the estimated accuracy of algorithms.

For this project, we will evaluate one linear, two non-linear, and three ensemble algorithms:

Linear Algorithm: Logistic Regression

Non-Linear Algorithms: Decision Trees (CART) and k-Nearest Neighbors

Ensemble Algorithms: Bagged CART, Random Forest, and Stochastic Gradient Boosting

The random number seed is reset before each run to ensure that the evaluation of each algorithm is performed using the same data splits. It ensures the results are directly comparable.

### 4.a) Generate models using linear algorithms

```{r LR}
# Logistic Regression (Classification)
email_notify(paste("Linear Regression modeling has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
fit.glm <- train(targetVar~., data=xy_train, method="glm", metric=metricTarget, trControl=control)
print(fit.glm)
proc.time()-startTimeModule
email_notify(paste("Linear Regression modeling completed!",date()))
```

### 4.b) Generate models using nonlinear algorithms

```{r CART}
# Decision Tree - CART (Regression/Classification)
email_notify(paste("Decision Tree modeling has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
fit.cart <- train(targetVar~., data=xy_train, method="rpart", metric=metricTarget, trControl=control)
print(fit.cart)
proc.time()-startTimeModule
email_notify(paste("Decision Tree modeling completed!",date()))
```

```{r KNN}
# k-Nearest Neighbors (Regression/Classification)
email_notify(paste("k-Nearest Neighbors modeling has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
fit.knn <- train(targetVar~., data=xy_train, method="knn", metric=metricTarget, trControl=control)
print(fit.knn)
proc.time()-startTimeModule
email_notify(paste("k-Nearest Neighbors modeling completed!",date()))
```

### 4.c) Generate models using ensemble algorithms
In this section, we will explore the use and tuning of ensemble algorithms to see whether we can improve the results.

```{r BAGCART}
# Bagged CART (Regression/Classification)
email_notify(paste("Bagged CART modeling has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
fit.bagcart <- train(targetVar~., data=xy_train, method="treebag", metric=metricTarget, trControl=control)
print(fit.bagcart)
proc.time()-startTimeModule
email_notify(paste("Bagged CART modeling completed!",date()))
```

```{r RF}
# Random Forest (Regression/Classification)
email_notify(paste("Random Forest modeling has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
fit.rf <- train(targetVar~., data=xy_train, method="rf", metric=metricTarget, trControl=control)
print(fit.rf)
proc.time()-startTimeModule
email_notify(paste("Random Forest modeling completed!",date()))
```

```{r GBM}
# Stochastic Gradient Boosting (Regression/Classification)
email_notify(paste("Stochastic Gradient Boosting modeling has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
fit.gbm <- train(targetVar~., data=xy_train, method="gbm", metric=metricTarget, trControl=control, verbose=F)
print(fit.gbm)
proc.time()-startTimeModule
email_notify(paste("Stochastic Gradient Boosting modeling completed!",date()))
```

### 4.d) Compare baseline algorithms

```{r SPOT_CHECK}
results <- resamples(list(LR=fit.glm, CART=fit.cart, kNN=fit.knn, BagCART=fit.bagcart, RF=fit.rf, GBM=fit.gbm))
summary(results)
dotplot(results)
cat('The average accuracy from all models is:',
    mean(c(results$values$`LR~Accuracy`,results$values$`CART~Accuracy`,results$values$`kNN~Accuracy`,results$values$`BagCART~Accuracy`,results$values$`RF~Accuracy`,results$values$`GBM~Accuracy`)))
```

## 5. Improve Accuracy or Results
After we achieve a short list of machine learning algorithms with good level of accuracy, we can leverage ways to improve the accuracy of the models.

Using the three best-perfoming algorithms from the previous section, we will Search for a combination of parameters for each algorithm that yields the best results.

### 5.a) Algorithm Tuning
Finally, we will tune the best-performing algorithms from each group further and see whether we can get more accuracy out of them.

```{r FINAL1}
# Tuning algorithm #1 - Stochastic Gradient Boosting
email_notify(paste("Algorithm #1 tuning has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
grid <- expand.grid(.n.trees=c(300,500,700,900), .shrinkage=0.1, .interaction.depth=3, .n.minobsinnode=10)
fit.final1 <- train(targetVar~., data=xy_train, method="gbm", metric=metricTarget, tuneGrid=grid, trControl=control, verbose=F)
print(fit.final1)
proc.time()-startTimeModule
email_notify(paste("Algorithm #1 tuning completed!",date()))
```

```{r FINAL2}
# Tuning algorithm #2 - Random Forest
email_notify(paste("Algorithm #2 tuning has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
grid <- expand.grid(mtry=c(10,20,30,40))
fit.final2 <- train(targetVar~., data=xy_train, method="rf", metric=metricTarget, tuneGrid=grid, trControl=control)
plot(fit.final2)
print(fit.final2)
proc.time()-startTimeModule
email_notify(paste("Algorithm #2 tuning completed!",date()))
```

### 5.d) Compare Algorithms After Tuning

```{r POST_TUNING}
results <- resamples(list(GBM=fit.final1, RF=fit.final2))
summary(results)
dotplot(results)
```

## 6. Finalize Model and Present Results
Once we have narrow down to a model that we believe can make accurate predictions on unseen data, we are ready to finalize it. Finalizing a model may involve sub-tasks such as:

* Using an optimal model tuned to make predictions on unseen data.
* Creating a standalone model using the tuned parameters
* Saving an optimal model to file for later use.

```{r}
email_notify(paste("Model Validation and Final Model Creation has begun!",date()))
```

### 6.a) Predictions on validation dataset

```{r PREDICT}
predictions <- predict(fit.final2, newdata=xy_test)
confusionMatrix(predictions, y_test)

pred <- prediction(as.numeric(predictions), as.numeric(y_test))
perf <- performance(pred, measure = "tpr", x.measure = "fpr")
plot(perf, colorize=TRUE)
auc <- performance(pred, measure = "auc")
auc <- auc@y.values[[1]]
auc
```

### 6.b) Create standalone model on entire training dataset

```{r FINALMODEL}
startTimeModule <- proc.time()
library(randomForest)
set.seed(seedNum)

# Combining the training and test datasets to form the complete dataset that will be used for training the final model
xy_complete <- rbind(xy_train, xy_test)

# Covert the target variable column to 0 and 1 for training the final GBM model
# names(xy_complete)[names(xy_complete)=='targetVar'] <- 'oldTarget'
# xy_complete$targetVar <- 0
# xy_complete$targetVar[xy_complete$oldTarget=="signal"] <- 1
# xy_complete$oldTarget <- NULL

finalModel <- randomForest(targetVar ~ ., data = xy_complete, mtry=8)
print(finalModel)
proc.time()-startTimeModule
```

### 6.c) Save model for later use

```{r}
#saveRDS(finalModel, "./finalModel_BinaryClass.rds")
```

```{r}
email_notify(paste("Model Validation and Final Model Creation Completed!",date()))
proc.time()-startTimeScript
```
