---
title: "Binary Classification Model for Seismic Bumps Take 1 Using R"
author: "David Lowe"
date: "August 29, 2018"
output: 
  html_document: 
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Template Credit: Adapted from a template made available by Dr. Jason Brownlee of Machine Learning Mastery. [http://machinelearningmastery.com/]

SUMMARY: The purpose of this project is to construct a prediction model using various machine learning algorithms and to document the end-to-end steps using a template. The Seismic Bumps Data Set is a binary-class classification situation where we are trying to predict one of the two possible outcomes.

INTRODUCTION: Mining activity has always been connected with the occurrence of dangers which are commonly called mining hazards. A special case of such a threat is a seismic hazard which frequently occurs in many underground mines. Seismic hazard is the hardest detectable and predictable of natural hazards, and it is comparable to an earthquake. The complexity of seismic processes and big disproportion between the number of low-energy seismic events and the number of high-energy phenomena causes the statistical techniques to be insufficient to predict seismic hazard. Therefore, it is essential to search for new opportunities for better hazard prediction, also using machine learning methods.

CONCLUSION: The baseline performance of the eight algorithms achieved an average accuracy of 93.11%. Three algorithms (Random Forest, Support Vector Machine, and Adaboost) achieved the top three accuracy scores after the first round of modeling. After a series of tuning trials, all three algorithms turned in the identical accuracy result of 93.42%, with an identical Kappa score of 0.0. With an imbalanced dataset we have on-hand, we will need to look for another metric or another approach to evaluate the models.

Dataset Used: Seismic Bumps Data Set

Dataset ML Model: Binary classification with numerical and categorical attributes

Dataset Reference: https://archive.ics.uci.edu/ml/datasets/seismic-bumps

The project aims to touch on the following areas:

1. Document a predictive modeling problem end-to-end.
2. Explore data cleaning and transformation options
3. Explore non-ensemble and ensemble algorithms for baseline model performance
4. Explore algorithm tuning techniques for improving model performance

Any predictive modeling machine learning project genrally can be broken down into about six major tasks:

1. Prepare Problem
2. Summarize Data
3. Prepare Data
4. Model and Evaluate Algorithms
5. Improve Accuracy or Results
6. Finalize Model and Present Results

## 1. Prepare Problem

### 1.a) Load libraries

```{r}
library(caret)
library(corrplot)
library(ROCR)
library(foreign)
```

### 1.b) Load dataset

```{r}
startTimeScript <- proc.time()
entireDataset <- read.arff("seismic-bumps.arff")

# Use variable totCol to hold the number of columns in the dataframe
totCol <- ncol(entireDataset)

# Rename the class/target column to a standard label
colnames(entireDataset)[totCol] <- "targetVar"
```

```{r}
# Setting up various variables to control the upcoming data visualization charts
# Set up a variable for the total number of attribute columns (totAttr)
totAttr <- totCol-1

#Which column does the first attribute column begin (1 - First, 2 - second)?
beginAttrCol <- 2
if (beginAttrCol == 1) {
attrColOffset <- 0
} else {
attrColOffset <- 1
}

# Set up the number of row and columns for visualization display. dispRow * dispCol should be >= totAttr
dispCol <- 4
if (totAttr%%dispCol == 0) {
dispRow <- totAttr%/%dispCol
} else {
dispRow <- (totAttr%/%dispCol) + 1
}
cat("Will attempt to create graphics grid (col x row): ", dispCol, ' by ', dispRow)
```

## 2. Summarize Data
To gain a better understanding of the data that we have on-hand, we will leverage a number of descriptive statistics and data visualization techniques. The plan is to use the results to consider new questions, review assumptions, and validate hypotheses that we can investigate later with specialized models.

### 2.a) Descriptive statistics

#### 2.a.i) Peek at the data itself.

```{r}
head(entireDataset)
```

#### 2.a.ii) Dimensions of the dataset.

```{r}
dim(entireDataset)
```

#### 2.a.iii) Types of the attributes.

```{r}
sapply(entireDataset, class)
```

#### 2.a.iv) Statistical summary of all attributes.

```{r}
summary(entireDataset)
```

#### 2.a.v) Summarize the levels of the class attribute.

```{r}
entireDataset_x <- entireDataset[,1:(totCol-1)]
entireDataset_y <- entireDataset[,totCol]
cbind(freq=table(entireDataset_y), percentage=prop.table(table(entireDataset_y))*100)
```

#### 2.a.vi) Count missing values.

```{r}
sapply(entireDataset, function(x) sum(is.na(x)))
```

### 2.b) Data visualizations

#### 2.b.i) Univariate plots to better understand each attribute.

```{r}
# Boxplots for each attribute
#par(mfrow=c(dispRow,dispCol))
#for(i in beginAttrCol:(totAttr+attrColOffset)) {
#	boxplot(entireDataset[,i], main=names(entireDataset)[i])
#}
```

```{r}
# Histograms each attribute
#par(mfrow=c(dispRow,dispCol))
#for(i in beginAttrCol:(totAttr+attrColOffset)) {
#	hist(entireDataset[,i], main=names(entireDataset)[i])
#}
```

```{r}
# Density plot for each attribute
#par(mfrow=c(dispRow,dispCol))
#for(i in beginAttrCol:(totAttr+attrColOffset)) {
#	plot(density(entireDataset[,i]), main=names(entireDataset)[i])
#}
```

#### 2.b.ii) Multivariate plots to better understand the relationships between attributes

```{r}
# Scatterplot matrix colored by class
#pairs(targetVar~., data=entireDataset, col=entireDataset$targetVar)
```

```{r}
# Box and whisker plots for each attribute by class
#scales <- list(x=list(relation="free"), y=list(relation="free"))
#featurePlot(x=entireDataset_x, y=entireDataset_y, plot="box", scales=scales)
```

```{r}
# Density plots for each attribute by class value
#featurePlot(x=entireDataset_x, y=entireDataset_y, plot="density", scales=scales)
```

```{r}
# Correlation plot
#correlations <- cor(entireDataset_x)
#corrplot(correlations, method="circle")
```

## 3. Prepare Data
Some dataset may require additional preparation activities that will best exposes the structure of the problem and the relationships between the input attributes and the output variable. Some data-prep tasks might include:

* Cleaning data by removing duplicates, marking missing values and even imputing missing values.
* Feature selection where redundant features may be removed.
* Data transforms where attributes are scaled or redistributed in order to best expose the structure of the problem later to learning algorithms.

### 3.a) Data Cleaning

```{r}
# Not applicable for this iteration of the project.

# Mark missing values
#invalid <- 0
#entireDataset$some_col[entireDataset$some_col==invalid] <- NA

# Impute missing values
#entireDataset$some_col <- with(entireDataset, impute(some_col, mean))
```

### 3.b) Feature Selection

```{r}
# Not applicable for this iteration of the project.
```

### 3.c) Data Transforms

```{r}
# Not applicable for this iteration of the project.
```

### 3.d) Split-out training and validation datasets
We create a training dataset (variable name "training") and a validation dataset (variable name "validation").

```{r}
seedNum <- 888
set.seed(seedNum)

# Create a list of the rows in the original dataset we can use for training
training_index <- createDataPartition(entireDataset$targetVar, p=0.70, list=FALSE)

# Use 70% of data to training and testing the models
training <- entireDataset[training_index,]

# Select the remaining 30% of the data for validation
validation <- entireDataset[-training_index,]
```

```{r}
proc.time()-startTimeScript
```

## 4. Model and Evaluate Algorithms
After the data-prep, we next work on finding a workable model by evaluating a subset of machine learning algorithms that are good at exploiting the structure of the training. The typical evaluation tasks include:

* Defining test options such as cross validation and the evaluation metric to use.
* Spot checking a suite of linear and nonlinear machine learning algorithms.
* Comparing the estimated accuracy of algorithms.

For this project, we will evaluate one linear, three non-linear, and four ensemble algorithms:

Linear Algorithm: Logistic Regression

Non-Linear Algorithms: Decision Trees (CART), k-Nearest Neighbors, and Support Vector Machine

Ensemble Algorithms: Bagged CART, Random Forest, Adaboost, and Stochastic Gradient Boosting

The random number seed is reset before each run to ensure that the evaluation of each algorithm is performed using the same data splits. It ensures the results are directly comparable.

### 4.a) Set test options and evaluation metric

```{r}
# Run algorithms using 10-fold cross validation
control <- trainControl(method="repeatedcv", number=10, repeats=3)
metricTarget <- "Accuracy"
```

### 4.b) Generate models using linear algorithms

```{r LR}
# Logistic Regression (Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.glm <- train(targetVar~., data=training, method="glm", metric=metricTarget, trControl=control)
proc.time()-startTimeModule
```

### 4.c) Generate models using nonlinear algorithms

```{r CART}
# Decision Tree - CART (Regression/Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.cart <- train(targetVar~., data=training, method="rpart", metric=metricTarget, trControl=control)
proc.time()-startTimeModule
```

```{r KNN}
# k-Nearest Neighbors (Regression/Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.knn <- train(targetVar~., data=training, method="knn", metric=metricTarget, trControl=control)
proc.time()-startTimeModule
```

```{r SVM}
# Support Vector Machine (Regression/Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.svm <- train(targetVar~., data=training, method="svmRadial", metric=metricTarget, trControl=control)
proc.time()-startTimeModule
```

### 4.d) Generate models using ensemble algorithms
In this section, we will explore the use and tuning of ensemble algorithms to see whether we can improve the results.

```{r BAGCART}
# Bagged CART (Regression/Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.bagcart <- train(targetVar~., data=training, method="treebag", metric=metricTarget, trControl=control)
proc.time()-startTimeModule
```

```{r RF}
# Random Forest (Regression/Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.rf <- train(targetVar~., data=training, method="rf", metric=metricTarget, trControl=control)
proc.time()-startTimeModule
```

```{r ADABOOST}
# AdaBoost (Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.ada <- train(targetVar~., data=training, method="ada", metric=metricTarget, trControl=control)
proc.time()-startTimeModule
```

```{r GBM}
# Stochastic Gradient Boosting (Regression/Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.gbm <- train(targetVar~., data=training, method="gbm", metric=metricTarget, trControl=control, verbose=F)
proc.time()-startTimeModule
```

### 4.e) Compare baseline algorithms

```{r SPOT_CHECK}
results <- resamples(list(LR=fit.glm, CART=fit.cart, kNN=fit.knn, SVM=fit.svm, BagCART=fit.bagcart, RF=fit.rf, AdaBoost=fit.ada, GBM=fit.gbm))
summary(results)
dotplot(results)
```

## 5. Improve Accuracy or Results
After we achieve a short list of machine learning algorithms with good level of accuracy, we can leverage ways to improve the accuracy of the models.

Using the three best-perfoming algorithms from the previous section, we will Search for a combination of parameters for each algorithm that yields the best results.

### 5.a) Algorithm Tuning
Finally, we will tune the best-performing algorithms from each group further and see whether we can get more accuracy out of them.

```{r FINAL1}
# Tuning algorithm #1 - Random Forest
startTimeModule <- proc.time()
set.seed(seedNum)
grid <- expand.grid(mtry = c(1:5))
fit.final1 <- train(targetVar~., data=training, method="rf", metric=metricTarget, tuneGrid=grid, trControl=control)
plot(fit.final1)
print(fit.final1)
proc.time()-startTimeModule
```

```{r FINAL2}
# Tuning algorithm #2 - Support Vector Machine
startTimeModule <- proc.time()
set.seed(seedNum)
grid <- expand.grid(sigma = c(.01, .015, 0.2), C = c(0.25, 0.5, 0.75, 1))
fit.final2 <- train(targetVar~., data=training, method="svmRadial", metric=metricTarget, tuneGrid=grid, trControl=control)
plot(fit.final2)
print(fit.final2)
proc.time()-startTimeModule
```

```{r FINAL3}
# Tuning algorithm #3 - Adaboost
startTimeModule <- proc.time()
set.seed(seedNum)
grid <- expand.grid(iter=c(50, 100, 150), maxdepth=c(1, 2, 3), nu=c(0.05, 0.1, 0.15))
fit.final3 <- train(targetVar~., data=training, method="ada", metric=metricTarget, tuneGrid=grid, trControl=control, verbose=F)
plot(fit.final3)
print(fit.final3)
proc.time()-startTimeModule
```

### 5.d) Compare Algorithms After Tuning

```{r POST_TUNING}
results <- resamples(list(RF=fit.final1, SVM=fit.final2, ADA=fit.final3))
summary(results)
dotplot(results)
```

## 6. Finalize Model and Present Results
Once we have narrow down to a model that we believe can make accurate predictions on unseen data, we are ready to finalize it. Finalizing a model may involve sub-tasks such as:

* Using an optimal model tuned to make predictions on unseen data.
* Creating a standalone model using the tuned parameters
* Saving an optimal model to file for later use.

### 6.a) Predictions on validation dataset

```{r PREDICT}
#predictions <- predict(fit.final2, newdata=validation)
#confusionMatrix(predictions, validation$targetVar)

#pred <- prediction(as.numeric(predictions), as.numeric(validation$targetVar))
#perf <- performance(pred, measure = "tpr", x.measure = "fpr")
#plot(perf, colorize=TRUE)
#auc <- performance(pred, measure = "auc")
#auc <- auc@y.values[[1]]
#auc
```

### 6.b) Create standalone model on entire training dataset

```{r FINALMODEL}
startTimeModule <- proc.time()
set.seed(seedNum)
#grid <- expand.grid(sigma = 0.015, C = 1)
#finalModel <- train(targetVar~., data=entireDataset, method="svmRadial", metric=metricTarget, tuneGrid=grid, trControl=control)
#print(finalModel)
proc.time()-startTimeModule
```

### 6.c) Save model for later use

```{r}
#saveRDS(finalModel, "./finalModel_BinaryClass.rds")
```

```{r}
proc.time()-startTimeScript
```
