---
title: "Binary Classification Model for Credit Card Default Using R Take 2"
author: "David Lowe"
date: "July 11, 2018"
output: 
  html_document: 
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Template Credit: Adapted from a template made available by Dr. Jason Brownlee of Machine Learning Mastery. [http://machinelearningmastery.com/]

Dataset Used: Default of Credit Card Clients Data Set

Dataset ML Model: Binary classification with numerical attributes

Dataset Reference: https://archive.ics.uci.edu/ml/datasets/default+of+credit+card+clients

One potential source of performance benchmark: https://www.kaggle.com/uciml/default-of-credit-card-clients-dataset

INTRODUCTION: This dataset contains information on default payments, demographic factors, credit data, history of payment, and bill statements of credit card clients in Taiwan from April 2005 to September 2005.

Previously on the Take No.1 iteration, the baseline performance of the ten algorithms achieved an average accuracy of 81.05%. Three algorithms (Support Vector Machine, AdaBoost, and Stochastic Gradient Boosting) achieved the top accuracy scores after the first round of modeling. After a series of tuning trials, Stochastic Gradient Boosting turned in the top result using the training data. It achieved an average accuracy of 82.18%. Using the optimized tuning parameter available, Stochastic Gradient Boosting algorithm processed the validation dataset with an accuracy of 81.94%, which was just slightly lower than the accuracy of the training data.

For the Take No.2 iteration, we will perform the binning operation for the credit limit and age attributes and observe the effects on the models.

CONCLUSION: The baseline performance of the ten algorithms achieved an average accuracy of 81.07%. Three algorithms (Decision Trees, AdaBoost, and Stochastic Gradient Boosting) achieved the top accuracy scores after the first round of modeling. After a series of tuning trials, AdaBoost turned in the top result using the training data. It achieved an average accuracy of 82.22%. Using the optimized tuning parameter available, the AdaBoost algorithm processed the validation dataset with an accuracy of 82.06%, which was just slightly lower than the accuracy of the training data. For this round of modeling, the Stochastic Gradient Boosting ensemble algorithm yielded consistently top-notch training and validation results, which warrant the additional processing required by the algorithm.

For this round of modeling, converting the credit limit and age attributes from ordinal to categorical did not have a noticeable effect on the accuracy of the models.

The purpose of this project is to analyze predictions using various machine learning algorithms and to document the steps using a template. Working through machine learning problems from end-to-end requires a structured modeling approach. Working problems through a project template can also encourage us to think about the problem more critically, to challenge our assumptions, and to get proficient at all parts of a modeling project.

The project aims to touch on the following areas:

1. Document a predictive modeling problem end-to-end.
2. Explore data cleaning and transformation options
3. Explore non-ensemble and ensemble algorithms for baseline model performance
4. Explore algorithm tuning techniques for improving model performance

Any predictive modeling machine learning project genrally can be broken down into about six major tasks:

1. Prepare Problem
2. Summarize Data
3. Prepare Data
4. Model and Evaluate Algorithms
5. Improve Accuracy or Results
6. Finalize Model and Present Results

## 1. Prepare Problem

### 1.a) Load libraries

```{r}
library(caret)
library(corrplot)
library(ROCR)
```

### 1.b) Load dataset

```{r}
startTimeScript <- proc.time()
entireDataset <- read.csv("default-of-credit-card-clients.csv", header= TRUE, stringsAsFactors = FALSE)

# Rename the class/target column to a standard label
colnames(entireDataset)[ncol(entireDataset)] <- "targetVar"

# Drop the Customer ID field as the label column has no relevance in the prediction exercise
entireDataset$CID <- NULL

# Use variable totCol to hold the number of columns in the dataframe
totCol <- ncol(entireDataset)
```

## 2. Summarize Data
To gain a better understanding of the data that we have on-hand, we will leverage a number of descriptive statistics and data visualization techniques. The plan is to use the results to consider new questions, review assumptions, and validate hypotheses that we can investigate later with specialized models.

### 2.a) Descriptive statistics

#### 2.a.i) Peek at the data itself.

```{r}
head(entireDataset)
```

#### 2.a.ii) Dimensions of the dataset.

```{r}
dim(entireDataset)
```

#### 2.a.iii) Types of the attributes.

```{r}
sapply(entireDataset, class)
```

#### 2.a.iv) Statistical summary of all attributes.

```{r}
summary(entireDataset)
```

#### 2.a.v) Summarize the levels of the class attribute.

```{r}
entireDataset_x <- entireDataset[,1:(totCol-1)]
entireDataset_y <- entireDataset[,totCol]
cbind(freq=table(entireDataset_y), percentage=prop.table(table(entireDataset_y))*100)
```

#### 2.a.vi) Count missing values.

```{r}
sapply(entireDataset, function(x) sum(is.na(x)))
```

### 2.b) Data visualizations

#### 2.b.i) Univariate plots to better understand each attribute.

```{r}
# Set up a variable for the total number of attribute columns (totAttr)
totAttr <- totCol-1
# Set up the number of row and columns for visualization display. dispRow * dispCol should be >= totAttr
dispCol <- 6
if (totAttr%%dispCol == 0) {
dispRow <- totAttr%/%dispCol
} else {
dispRow <- (totAttr%/%dispCol) + 1
}
cat("Will attempt to create graphics grid (col x row): ", dispCol, ' by ', dispRow)
```


```{r}
# Boxplots for each attribute
par(mfrow=c(dispRow,dispCol))
for(i in 1:totAttr) {
	boxplot(entireDataset[,i], main=names(entireDataset)[i])
}
```

```{r}
# Histograms each attribute
par(mfrow=c(dispRow,dispCol))
for(i in 1:totAttr) {
	hist(entireDataset[,i], main=names(entireDataset)[i])
}
```

```{r}
# Density plot for each attribute
par(mfrow=c(dispRow,dispCol))
for(i in 1:totAttr) {
	plot(density(entireDataset[,i]), main=names(entireDataset)[i])
}
```

#### 2.b.ii) Multivariate plots to better understand the relationships between attributes

```{r}
# Scatterplot matrix colored by class
pairs(targetVar~., data=entireDataset, col=entireDataset$targetVar)
```

```{r}
# Box and whisker plots for each attribute by class
scales <- list(x=list(relation="free"), y=list(relation="free"))
featurePlot(x=entireDataset_x, y=entireDataset_y, plot="box", scales=scales)
```

```{r}
# Density plots for each attribute by class value
featurePlot(x=entireDataset_x, y=entireDataset_y, plot="density", scales=scales)
```

```{r}
# Correlation plot
correlations <- cor(entireDataset_x)
corrplot(correlations, method="circle")
```

## 3. Prepare Data
Some dataset may require additional preparation activities that will best exposes the structure of the problem and the relationships between the input attributes and the output variable. Some data-prep tasks might include:

* Cleaning data by removing duplicates, marking missing values and even imputing missing values.
* Feature selection where redundant features may be removed.
* Data transforms where attributes are scaled or redistributed in order to best expose the structure of the problem later to learning algorithms.

### 3.a) Data Cleaning

```{r}
# Correct the invalid values in the Education and Marital Status columns
entireDataset$EDUCATION[entireDataset$EDUCATION==0] <- 4
entireDataset$EDUCATION[entireDataset$EDUCATION==5] <- 4
entireDataset$EDUCATION[entireDataset$EDUCATION==6] <- 4
entireDataset$MARRIAGE[entireDataset$MARRIAGE==0] <- 3

# Impute missing values
#entireDataset$some_col <- with(entireDataset, impute(some_col, mean))
```

### 3.b) Feature Selection

```{r}
# Not applicable for this iteration of the project.
```

### 3.c) Data Transforms

```{r}
# Conver the integer variables to categorical variables as appropriate
entireDataset$SEX <- as.factor(entireDataset$SEX)
entireDataset$EDUCATION <- as.factor(entireDataset$EDUCATION)
entireDataset$MARRIAGE <- as.factor(entireDataset$MARRIAGE)
entireDataset$targetVar <- as.factor(entireDataset$targetVar)

# Conver the credit limit attribute into bins
credit_bins <- c(0, 50000, 100000, 200000, 500000, 800000, 1000000)
credit_labels <- c("50K", "100K", "200K", "500K", "800K", "1M")
entireDataset$CREDIT_BIN <- cut(entireDataset$LIMIT_BAL, breaks = credit_bins, labels = credit_labels)

# Drop the Credit Limit field after binning
entireDataset$LIMIT_BAL <- NULL

# Conver the age attribute into bins
age_bins <- c(20, 30, 40, 50, 60, 70, 80)
age_labels <- c("20s", "30s", "40s", "50s", "60s", "70s")
entireDataset$AGE_BIN <- cut(entireDataset$AGE, breaks = age_bins, labels = age_labels)

# Drop the Age field after binning
entireDataset$AGE <- NULL
```

### 3.d) Split-out training and validation datasets
We create a training dataset (variable name "training") and a validation dataset (variable name "validation").

```{r}
seedNum <- 888
set.seed(seedNum)

# Create a list of 70% of the rows in the original dataset we can use for training
training_index <- createDataPartition(entireDataset$targetVar, p=0.70, list=FALSE)

# Use 70% of data to training and testing the models
training <- entireDataset[training_index,]

# Select the remaining 30% of the data for validation
validation <- entireDataset[-training_index,]
```

```{r}
proc.time()-startTimeScript
```

## 4. Model and Evaluate Algorithms
After the data-prep, we next work on finding a workable model by evaluating a subset of machine learning algorithms that are good at exploiting the structure of the training. The typical evaluation tasks include:

* Defining test options such as cross validation and the evaluation metric to use.
* Spot checking a suite of linear and nonlinear machine learning algorithms.
* Comparing the estimated accuracy of algorithms.

For this project, we will evaluate one linear, three non-linear, and four ensemble algorithms:

Linear Algorithm: Logistic Regression

Non-Linear Algorithms: Decision Trees (CART), k-Nearest Neighbors, and Support Vector Machine

Ensemble Algorithms: Bagged CART, Random Forest, Adaboost, and Stochastic Gradient Boosting

The random number seed is reset before each run to ensure that the evaluation of each algorithm is performed using the same data splits. It ensures the results are directly comparable.

### 4.a) Set test options and evaluation metric

```{r}
# Run algorithms using 10-fold cross validation
control <- trainControl(method="repeatedcv", number=10, repeats=3)
metricTarget <- "Accuracy"
```

### 4.b) Generate models using linear algorithms

```{r LR}
# Logistic Regression (Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.glm <- train(targetVar~., data=training, method="glm", metric=metricTarget, trControl=control)
proc.time()-startTimeModule
```

### 4.c) Generate models using nonlinear algorithms

```{r CART}
# Decision Tree - CART (Regression/Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.cart <- train(targetVar~., data=training, method="rpart", metric=metricTarget, trControl=control)
proc.time()-startTimeModule
```

```{r KNN}
# k-Nearest Neighbors (Regression/Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.knn <- train(targetVar~., data=training, method="knn", metric=metricTarget, trControl=control)
proc.time()-startTimeModule
```

```{r SVM}
# Support Vector Machine (Regression/Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.svm <- train(targetVar~., data=training, method="svmRadial", metric=metricTarget, trControl=control)
proc.time()-startTimeModule
```

### 4.d) Generate models using ensemble algorithms
In this section, we will explore the use and tuning of ensemble algorithms to see whether we can improve the results.

```{r BAGCART}
# Bagged CART (Regression/Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.bagcart <- train(targetVar~., data=training, method="treebag", metric=metricTarget, trControl=control)
proc.time()-startTimeModule
```

```{r RF}
# Random Forest (Regression/Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.rf <- train(targetVar~., data=training, method="rf", metric=metricTarget, trControl=control)
proc.time()-startTimeModule
```

```{r ADABOOST}
# AdaBoost (Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.ada <- train(targetVar~., data=training, method="ada", metric=metricTarget, trControl=control)
proc.time()-startTimeModule
```

```{r GBM}
# Stochastic Gradient Boosting (Regression/Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.gbm <- train(targetVar~., data=training, method="gbm", metric=metricTarget, trControl=control, verbose=F)
proc.time()-startTimeModule
```

### 4.e) Compare baseline algorithms

```{r SPOT_CHECK}
results <- resamples(list(LR=fit.glm, CART=fit.cart, kNN=fit.knn, SVM=fit.svm, BagCART=fit.bagcart, RF=fit.rf, AdaBoost=fit.ada, GBM=fit.gbm))
summary(results)
dotplot(results)
```

## 5. Improve Accuracy or Results
After we achieve a short list of machine learning algorithms with good level of accuracy, we can leverage ways to improve the accuracy of the models.

Using the three best-perfoming algorithms from the previous section, we will Search for a combination of parameters for each algorithm that yields the best results.

### 5.a) Algorithm Tuning
Finally, we will tune the best-performing algorithms from each group further and see whether we can get more accuracy out of them.

```{r FINAL1}
# Tuning algorithm #1 - AdaBoost
startTimeModule <- proc.time()
set.seed(seedNum)
grid <- expand.grid(iter = c(50, 100, 150), maxdepth = c(1, 2, 3), nu = c(0.05, 0.1, 0.2))
fit.final1 <- train(targetVar~., data=training, method="ada", metric=metricTarget, tuneGrid=grid, trControl=control)
print(fit.final1)
proc.time()-startTimeModule
```

```{r FINAL2}
# Tuning algorithm #2 - Decision Tree
startTimeModule <- proc.time()
set.seed(seedNum)
grid <- expand.grid(cp = c(0.1, 0.5, 0.9))
fit.final2 <- train(targetVar~., data=training, method="rpart", metric=metricTarget, tuneGrid=grid, trControl=control)
plot(fit.final2)
print(fit.final2)
proc.time()-startTimeModule
```

```{r FINAL3}
# Tuning algorithm #3 - Stochastic Gradient Boosting
startTimeModule <- proc.time()
set.seed(seedNum)
grid <- expand.grid(.n.trees=c(50, 100, 200), .shrinkage=c(0.01, 0.05, 0.1), .interaction.depth=c(1, 2, 3), .n.minobsinnode=c(3, 5, 7))
fit.final3 <- train(targetVar~., data=training, method="gbm", metric=metricTarget, tuneGrid=grid, trControl=control, verbose=F)
plot(fit.final3)
print(fit.final3)
proc.time()-startTimeModule
```

### 5.d) Compare Algorithms After Tuning

```{r POST_TUNING}
results <- resamples(list(ADA=fit.final1, CART=fit.final2, GBM=fit.final3))
summary(results)
dotplot(results)
```

## 6. Finalize Model and Present Results
Once we have narrow down to a model that we believe can make accurate predictions on unseen data, we are ready to finalize it. Finalizing a model may involve sub-tasks such as:

* Using an optimal model tuned to make predictions on unseen data.
* Creating a standalone model using the tuned parameters
* Saving an optimal model to file for later use.

### 6.a) Predictions on validation dataset

```{r PREDICT}
predictions <- predict(fit.final1, newdata=validation)
confusionMatrix(predictions, validation$targetVar)

pred <- prediction(as.numeric(predictions), as.numeric(validation$targetVar))
perf <- performance(pred, measure = "tpr", x.measure = "fpr")
plot(perf, colorize=TRUE)
auc <- performance(pred, measure = "auc")
auc <- auc@y.values[[1]]
auc
```

### 6.b) Create standalone model on entire training dataset

```{r FINALMODEL}
startTimeModule <- proc.time()
set.seed(seedNum)
grid <- expand.grid(iter = 100, maxdepth = 3, nu = 0.05)
finalModel <- train(targetVar~., data=entireDataset, method="ada", metric=metricTarget, tuneGrid=grid, trControl=control, verbose=F)
print(finalModel)
proc.time()-startTimeModule
```

### 6.c) Save model for later use

```{r}
#saveRDS(finalModel, "./finalModel_BinaryClass.rds")
```

```{r}
proc.time()-startTimeScript
```
