---
title: "Binary Classification Model for Heart Disease Study Using R Take 4"
author: "David Lowe"
date: "May 30, 2019"
output: 
  html_document: 
    toc: yes
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Template Credit: Adapted from a template made available by Dr. Jason Brownlee of Machine Learning Mastery. [http://machinelearningmastery.com/]

SUMMARY: The purpose of this project is to construct a prediction model using various machine learning algorithms and to document the end-to-end steps using a template. The Heart Disease dataset is a binary classification situation where we are trying to predict one of the two possible outcomes.

INTRODUCTION: The original database contains 76 attributes, but all published experiments refer to using a subset of 14 of them. In particular, the Cleveland database is the only one that has been used by machine learning researchers to this date. The "num" field refers to the presence of heart disease in the patient. It is integer valued from 0 (no presence) to 4. Experiments with the Cleveland database have concentrated on simply attempting to distinguish presence (values 1,2,3,4) from absence (value 0).

In iteration Take1, we examined the Cleveland dataset and create a machine learning model to fit the data.

In iteration Take2, we examined the Hungarian dataset and create a machine learning model to fit the data.

In iteration Take3, we examined the Switzerland dataset and create a machine learning model to fit the data.

In this iteration, we will examine the Long Beach VA dataset and create a machine learning model to fit the data.

ANALYSIS: The baseline performance of the machine learning algorithms achieved an average accuracy of 73.74%. Two algorithms (Random Forest and Gradient Boosting) achieved the top accuracy metrics after the first round of modeling. After a series of tuning trials, Random Forest turned in the top overall result and achieved an accuracy metric of 74.71%. By using the optimized parameters, the Random Forest algorithm processed the testing dataset with an accuracy of 79.66%, which was even better than the prediction accuracy gained from the training data.

CONCLUSION: For the Long Beach VA dataset, the Random Forest algorithm achieved the best overall results using the training and testing datasets. For this dataset, Random Forest could be considered for further modeling or production use.

Dataset Used: Heart Disease Data Set

Dataset ML Model: Binary classification with numerical and categorical attributes

Dataset Reference: https://archive.ics.uci.edu/ml/datasets/Heart+Disease

One potential source of performance benchmark: https://www.kaggle.com/ronitf/heart-disease-uci

The project aims to touch on the following areas:

1. Document a predictive modeling problem end-to-end.
2. Explore data cleaning and transformation options
3. Explore non-ensemble and ensemble algorithms for baseline model performance
4. Explore algorithm tuning techniques for improving model performance

Any predictive modeling machine learning project genrally can be broken down into about six major tasks:

1. Prepare Problem
2. Summarize Data
3. Prepare Data
4. Model and Evaluate Algorithms
5. Improve Accuracy or Results
6. Finalize Model and Present Results

## 1. Prepare Problem

### 1.a) Load libraries

```{r}
startTimeScript <- proc.time()
library(caret)
library(corrplot)
library(mailR)
library(stringr)
library(MLmetrics)
library(DMwR)
library(ROCR)
library(mlbench)
library(Hmisc)

# Create one random seed number for reproducible results
seedNum <- 888
set.seed(seedNum)
```

### 1.b) Set up the email notification function

```{r}
email_notify <- function(msg=""){
  sender <- Sys.getenv("MAIL_SENDER")
  receiver <- Sys.getenv("MAIL_RECEIVER")
  gateway <- Sys.getenv("SMTP_GATEWAY")
  smtpuser <- Sys.getenv("SMTP_USERNAME")
  password <- Sys.getenv("SMTP_PASSWORD")
  sbj_line <- "Notification from R Binary Classification Script"
  send.mail(
    from = sender,
    to = receiver,
    subject= sbj_line,
    body = msg,
    smtp = list(host.name = gateway, port = 587, user.name = smtpuser, passwd = password, ssl = TRUE),
    authenticate = TRUE,
    send = TRUE)
}
```

```{r}
email_notify(paste("Library and Data Loading has begun!",date()))
```

### 1.c) Load dataset

```{r}
# Slicing up the document path to get the final destination file name
dataset_path <- 'https://archive.ics.uci.edu/ml/machine-learning-databases/heart-disease/processed.va.data'
doc_path_list <- str_split(dataset_path, "/")
dest_file <- doc_path_list[[1]][length(doc_path_list[[1]])]

if (!file.exists(dest_file)) {
  # Download the document from the website
  cat("Downloading", dataset_path, "as", dest_file, "\n")
  download.file(dataset_path, dest_file, mode = "wb")
  cat(dest_file, "downloaded!\n")
#  unzip(dest_file)
#  cat(dest_file, "unpacked!\n")
} else {
  cat(dest_file, 'already exists. Will not download again!')
}

inputFile <- dest_file
colNames <- c('age','sex','cp','trestbps','chol','fbs','restecg','thalach','exang','oldpeak','slope','ca','thal','num')
xy_original <- read.csv(inputFile, sep = ',', header = FALSE, col.names = colNames, na.strings = "?")
```

```{r}
# Take a peek at the dataframe after the import
head(xy_original)
```

```{r}
sapply(xy_original, class)
```

```{r}
sapply(xy_original, function(x) sum(is.na(x)))
```

### 1.d) Data Cleaning

```{r}
# Due to the large number of missing data in these columns, we will drop them from the dataset
xy_original$ca <- NULL
xy_original$thal <- NULL

# Perform some immediate cleaning tasks to get data into the right formats and types
xy_original$sex <- as.factor(xy_original$sex)
xy_original$cp <- as.factor(xy_original$cp)
xy_original$restecg <- as.factor(xy_original$restecg)

# impute the missing values with values appropriate for the attributes
xy_original$trestbps <- with(xy_original, impute(trestbps, median))

chol_median <- median(xy_original$chol, na.rm = TRUE)
xy_original$chol[xy_original$chol==0] <- chol_median
xy_original$chol <- with(xy_original, impute(chol, chol_median))

xy_original$fbs <- with(xy_original, impute(fbs, 0))
xy_original$fbs <- as.factor(xy_original$fbs)

xy_original$thalach <- with(xy_original, impute(thalach, median))

xy_original$exang <- with(xy_original, impute(exang, 0))
xy_original$exang <- as.factor(xy_original$exang)

xy_original$oldpeak <- with(xy_original, impute(oldpeak, median))

xy_original$slope <- with(xy_original, impute(slope, 0))
xy_original$slope <- as.factor(xy_original$slope)
```

```{r}
# Transform the num codes to their corresponding categories
reClassNum <- function(num_code) {
  num_category <-
    if (num_code >= 1) 1 else 0
  return(num_category)
}

xy_original$num <- sapply(xy_original$num, reClassNum)
xy_original$num <- as.factor(xy_original$num)
```

```{r}
# Take a peek at the dataframe after the cleaning
head(xy_original)
```

```{r}
sapply(xy_original, class)
```

```{r}
sapply(xy_original, function(x) sum(is.na(x)))
```

### 1.e) Splitting Data into Training and Testing Sets

```{r}
# Use variable totCol to hold the number of columns in the dataframe
totCol <- ncol(xy_original)

# Set up variable totAttr for the total number of attribute columns
totAttr <- totCol-1
```

```{r}
# targetCol variable indicates the column location of the target/class variable
# If the first column, set targetCol to 1. If the last column, set targetCol to totCol
# if (targetCol <> 1) and (targetCol <> totCol), be aware when slicing up the dataframes for visualization! 
targetCol <- totCol

# Standardize the class column to the name of targetVar if applicable
colnames(xy_original)[targetCol] <- "targetVar"
#xy_original$targetVar <- relevel(xy_original$targetVar,"pos")
```

```{r}
# We create training datasets (xy_train, x_train, y_train) for various visualization and cleaning/transformation operations.
# We create testing datasets (xy_test, y_test) for various visualization and cleaning/transformation operations.
set.seed(seedNum)

# Create a list of the rows in the original dataset we can use for training
# Use 70% of the data to train the models and the remaining for testing/validation
training_index <- createDataPartition(xy_original$targetVar, p=0.70, list=FALSE)
xy_train <- xy_original[training_index,]
xy_test <- xy_original[-training_index,]

if (targetCol==1) {
x_train <- xy_train[,(targetCol+1):totCol]
y_train <- xy_train[,targetCol]
y_test <- xy_test[,targetCol]
} else {
x_train <- xy_train[,1:(totAttr)]
y_train <- xy_train[,totCol]
y_test <- xy_test[,totCol]
}
```

### 1.f) Set up the key parameters to be used in the script

```{r}
# Set up the number of row and columns for visualization display. dispRow * dispCol should be >= totAttr
dispCol <- 4
if (totAttr%%dispCol == 0) {
dispRow <- totAttr%/%dispCol
} else {
dispRow <- (totAttr%/%dispCol) + 1
}
cat("Will attempt to create graphics grid (col x row): ", dispCol, ' by ', dispRow)
```

```{r}
# Run algorithms using 10-fold cross validation
control <- trainControl(method="repeatedcv", number=10, repeats=3)
metricTarget <- "Accuracy"
```

```{r}
email_notify(paste("Library and Data Loading completed!",date()))
```

## 2. Summarize Data
To gain a better understanding of the data that we have on-hand, we will leverage a number of descriptive statistics and data visualization techniques. The plan is to use the results to consider new questions, review assumptions, and validate hypotheses that we can investigate later with specialized models.

```{r}
email_notify(paste("Data Summarization and Visualization has begun!",date()))
```

### 2.a) Descriptive statistics

#### 2.a.i) Peek at the data itself.

```{r}
head(xy_train)
```

#### 2.a.ii) Dimensions of the dataset.

```{r}
dim(xy_train)
```

#### 2.a.iii) Types of the attributes.

```{r}
sapply(xy_train, class)
```

#### 2.a.iv) Statistical summary of all attributes.

```{r}
summary(xy_train)
```

#### 2.a.v) Count missing values.

```{r}
sapply(xy_train, function(x) sum(is.na(x)))
```

#### 2.a.vi) Summarize the levels of the class attribute.

```{r}
cbind(freq=table(y_train), percentage=prop.table(table(y_train))*100)
```

### 2.b) Data visualizations

#### 2.b.i) Univariate plots to better understand each attribute.

```{r}
# Boxplots for each attribute
# par(mfrow=c(dispRow,dispCol))
# for(i in 1:totAttr) {
# 	boxplot(x_train[,i], main=names(x_train)[i])
# }
```

```{r}
# Histograms each attribute
# par(mfrow=c(dispRow,dispCol))
# for(i in 1:totAttr) {
# 	hist(x_train[,i], main=names(x_train)[i])
# }
```

```{r}
# Density plot for each attribute
# par(mfrow=c(dispRow,dispCol))
# for(i in 1:totAttr) {
# 	plot(density(x_train[,i]), main=names(x_train)[i])
# }
```

#### 2.b.ii) Multivariate plots to better understand the relationships between attributes

```{r}
# Scatterplot matrix colored by class
#pairs(targetVar~., data=xy_train, col=xy_train$targetVar)
```

```{r}
# Box and whisker plots for each attribute by class
#scales <- list(x=list(relation="free"), y=list(relation="free"))
#featurePlot(x=x_train, y=y_train, plot="box", scales=scales)
```

```{r}
# Density plots for each attribute by class value
#featurePlot(x=x_train, y=y_train, plot="density", scales=scales)
```

```{r}
# Correlation plot
# correlations <- cor(x_train)
# corrplot(correlations, method="circle")
```

```{r}
email_notify(paste("Data Summarization and Visualization completed!",date()))
```

## 3. Prepare Data
Some dataset may require additional preparation activities that will best exposes the structure of the problem and the relationships between the input attributes and the output variable. Some data-prep tasks might include:

* Cleaning data by removing duplicates, marking missing values and even imputing missing values.
* Feature selection where redundant features may be removed.
* Data transforms where attributes are scaled or redistributed in order to best expose the structure of the problem later to learning algorithms.

```{r}
email_notify(paste("Data Cleaning and Transformation has begun!",date()))
```

### 3.a) Feature Selection

```{r}
# Not applicable for this iteration of the project
```

### 3.b) Data Transforms

```{r}
# Not applicable for this iteration of the project
```

### 3.c) Display the Final Dataset for Model-Building

```{r}
dim(xy_train)
dim(xy_test)
```

```{r}
sapply(xy_train, class)
```

```{r}
email_notify(paste("Data Cleaning and Transformation completed!",date()))
proc.time()-startTimeScript
```

## 4. Model and Evaluate Algorithms
After the data-prep, we next work on finding a workable model by evaluating a subset of machine learning algorithms that are good at exploiting the structure of the training. The typical evaluation tasks include:

* Defining test options such as cross validation and the evaluation metric to use.
* Spot checking a suite of linear and nonlinear machine learning algorithms.
* Comparing the estimated accuracy of algorithms.

For this project, we will evaluate one linear, one non-linear, and three ensemble algorithms:

Linear Algorithm: Logistic Regression

Non-Linear Algorithm: Decision Trees (CART)

Ensemble Algorithms: Bagged CART, Random Forest, and Gradient Boosting

The random number seed is reset before each run to ensure that the evaluation of each algorithm is performed using the same data splits. It ensures the results are directly comparable.

### 4.a) Generate models using linear algorithms

```{r LR}
# Logistic Regression (Classification)
email_notify(paste("Logistic Regression modeling has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
fit.glm <- train(targetVar~., data=xy_train, method="glm", metric=metricTarget, trControl=control)
print(fit.glm)
proc.time()-startTimeModule
email_notify(paste("Logistic Regression modeling completed!",date()))
```

### 4.b) Generate models using nonlinear algorithms

```{r CART}
# Decision Tree - CART (Regression/Classification)
email_notify(paste("Decision Tree modeling has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
fit.cart <- train(targetVar~., data=xy_train, method="rpart", metric=metricTarget, trControl=control)
print(fit.cart)
proc.time()-startTimeModule
email_notify(paste("Decision Tree modeling completed!",date()))
```

### 4.c) Generate models using ensemble algorithms
In this section, we will explore the use and tuning of ensemble algorithms to see whether we can improve the results.

```{r BAGCART}
# Bagged CART (Regression/Classification)
email_notify(paste("Bagged CART modeling has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
fit.bagcart <- train(targetVar~., data=xy_train, method="treebag", metric=metricTarget, trControl=control)
print(fit.bagcart)
proc.time()-startTimeModule
email_notify(paste("Bagged CART modeling completed!",date()))
```

```{r RF}
# Random Forest (Regression/Classification)
email_notify(paste("Random Forest modeling has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
fit.rf <- train(targetVar~., data=xy_train, method="rf", metric=metricTarget, trControl=control)
print(fit.rf)
proc.time()-startTimeModule
email_notify(paste("Random Forest modeling completed!",date()))
```

```{r GBM}
# Gradient Boosting (Regression/Classification)
email_notify(paste("Gradient Boosting modeling has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
fit.gbm <- train(targetVar~., data=xy_train, method="xgbTree", metric=metricTarget, trControl=control, verbose=F)
# fit.gbm <- train(targetVar~., data=xy_train, method="gbm", metric=metricTarget, trControl=control, verbose=F)
print(fit.gbm)
proc.time()-startTimeModule
email_notify(paste("Gradient Boosting modeling completed!",date()))
```

### 4.d) Compare baseline algorithms

```{r SPOT_CHECK}
results <- resamples(list(LR=fit.glm, CART=fit.cart, BagCART=fit.bagcart, RF=fit.rf, GBM=fit.gbm))
summary(results)
dotplot(results)
cat('The average accuracy from all models is:',
    mean(c(results$values$`LR~Accuracy`,results$values$`CART~Accuracy`,results$values$`BagCART~Accuracy`,results$values$`RF~Accuracy`,results$values$`GBM~Accuracy`)))
```

## 5. Improve Accuracy or Results
After we achieve a short list of machine learning algorithms with good level of accuracy, we can leverage ways to improve the accuracy of the models.

Using the three best-perfoming algorithms from the previous section, we will Search for a combination of parameters for each algorithm that yields the best results.

### 5.a) Algorithm Tuning
Finally, we will tune the best-performing algorithms from each group further and see whether we can get more accuracy out of them.

```{r FINAL1}
# Tuning algorithm #1 - Random Forest
email_notify(paste("Algorithm #1 tuning has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
grid <- expand.grid(mtry=c(2,5,8,11))
fit.final1 <- train(targetVar~., data=xy_train, method="rf", metric=metricTarget, tuneGrid=grid, trControl=control)
plot(fit.final1)
print(fit.final1)
proc.time()-startTimeModule
email_notify(paste("Algorithm #1 tuning completed!",date()))
```

```{r FINAL2}
# Tuning algorithm #2 - Gradient Boosting
email_notify(paste("Algorithm #2 tuning has begun!",date()))
startTimeModule <- proc.time()
set.seed(seedNum)
grid <- expand.grid(nrounds=c(10,25,50,100), max_depth=1, eta=0.3, gamma=0, colsample_bytree=0.8, min_child_weight=1, subsample=0.75)
fit.final2 <- train(targetVar~., data=xy_train, method="xgbTree", metric=metricTarget, tuneGrid=grid, trControl=control, verbose=F)
plot(fit.final2)
print(fit.final2)
proc.time()-startTimeModule
email_notify(paste("Algorithm #2 tuning completed!",date()))
```

### 5.d) Compare Algorithms After Tuning

```{r POST_TUNING}
results <- resamples(list(RF=fit.final1, GBM=fit.final2))
summary(results)
dotplot(results)
```

## 6. Finalize Model and Present Results
Once we have narrow down to a model that we believe can make accurate predictions on unseen data, we are ready to finalize it. Finalizing a model may involve sub-tasks such as:

* Using an optimal model tuned to make predictions on unseen data.
* Creating a standalone model using the tuned parameters
* Saving an optimal model to file for later use.

```{r}
email_notify(paste("Model Validation and Final Model Creation has begun!",date()))
```

### 6.a) Predictions on validation dataset

```{r PREDICT}
predictions <- predict(fit.final1, newdata=xy_test)
confusionMatrix(predictions, y_test)

pred <- prediction(as.numeric(predictions), as.numeric(y_test))
perf <- performance(pred, measure = "tpr", x.measure = "fpr")
plot(perf, colorize=TRUE)
auc <- performance(pred, measure = "auc")
auc <- auc@y.values[[1]]
auc
```

### 6.b) Create standalone model on entire training dataset

```{r FINALMODEL}
startTimeModule <- proc.time()
library(xgboost)
set.seed(seedNum)

# Combining the training and test datasets to form the original dataset that will be used for training the final model
xy_complete <- rbind(xy_train, xy_test)
y_final <- xy_complete$targetVar
xy_complete$targetVar <- NULL
x_final <- as.matrix(xy_complete)

# finalModel <- xgboost(data=x_final, label=y_final, nrounds=10, max_depth=1, eta=0.3, gamma=0, colsample_bytree=0.8, min_child_weight=1, subsample=0.5)
# print(finalModel)
proc.time()-startTimeModule
```

### 6.c) Save model for later use

```{r}
#saveRDS(finalModel, "./finalModel_BinaryClass.rds")
```

```{r}
email_notify(paste("Model Validation and Final Model Creation Completed!",date()))
proc.time()-startTimeScript
```
