---
title: "Simple Classification Model for Bank Marketing Using R"
author: "David Lowe"
date: "April 20, 2018"
output: 
  html_document: 
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Template Credit: Adapted from a template made available by Dr. Jason Brownlee of Machine Learning Mastery (http://machinelearningmastery.com/)

Dataset Used: Bank Marketing Data Set

Data Set ML Model: Binary classification with numerical and categorical attributes

Dataset Reference: http://archive.ics.uci.edu/ml/datasets/bank+marketing

One source of potential performance benchmarks: https://www.kaggle.com/rouseguy/bankbalanced

INTRODUCTION: The Bank Marketing dataset involves predicting the whether the bank clients will subscribe (yes/no) a term deposit (target variable). It is a binary (2-class) classification problem. There are over 45,000 observations with 16 input variables and 1 output variable. There are no missing values within the dataset.

CONCLUSION: The baseline performance of eight algorithms achieved an average accuracy of 89.99%. Three algorithms (Random Forest, Stochastic Gradient Boosting, and Bagged CART) achieved the top accuracy and Kappa scores. The top result achieved using the training data was from Random Forest. It achieved an average accuracy of 90.65% after a series of tuning trials, and its accuracy on processing the validation dataset was 90.91%. For this project, the Random Forest ensemble algorithms yielded consistently top-notch training and validation results, which warrant the additional processing required by the algorithm.

The purpose of this project is to analyze predictions using various machine learning algorithms and to document the steps using a template. Working through machine learning problems from end-to-end requires a structured modeling approach. Working problems through a project template can also encourage us to think about the problem more critically, to challenge our assumptions, and to get proficient at all parts of a modeling project.

The project aims to touch on the following areas:

1. Document a predictive modeling problem end-to-end.
2. Explore data cleaning and transformation options
3. Explore non-ensemble and ensemble algorithms for baseline model performance
4. Explore algorithm tuning techniques for improving model performance

Any predictive modeling machine learning project genrally can be broken down into about six major tasks:

1. Prepare Problem
2. Summarize Data
3. Prepare Data
4. Model and Evaluate Algorithms
5. Improve Accuracy or Results
6. Finalize Model and Present Results

## 1. Prepare Problem

### 1.a) Load libraries

```{r}
library(caret)
library(ROCR)
```

### 1.b) Load training

```{r}
entireDataset <- read.csv2("bank-full.csv", header= TRUE)

# Use variable totCol to hold the number of columns in the dataframe
totCol <- ncol(entireDataset)

# Rename the class column to a standard label
colnames(entireDataset)[ncol(entireDataset)] <- "targetVar"
```

## 2. Summarize Data
To gain a better understanding of the data that we have on-hand, we will leverage a number of descriptive statistics and data visualization techniques. The plan is to use the results to consider new questions, review assumptions, and validate hypotheses that we can investigate later with specialized models.

### 2.a) Descriptive statistics

#### 2.a.i) Peek at the data itself.

```{r}
head(entireDataset)
```

#### 2.a.ii) Dimensions of the dataset.

```{r}
dim(entireDataset)
```

#### 2.a.iii) Types of the attributes.

```{r}
sapply(entireDataset, class)
```

#### 2.a.iv) Statistical summary of all attributes.

```{r}
summary(entireDataset)
```

#### 2.a.v) Summarize the levels of the class attribute.

```{r}
entireDataset_x <- entireDataset[,1:(totCol-1)]
entireDataset_y <- entireDataset[,totCol]
cbind(freq=table(entireDataset_y), percentage=prop.table(table(entireDataset_y))*100)
```

#### 2.a.vi) Count missing values.

```{r}
sapply(entireDataset, function(x) sum(is.na(x)))
```

### 2.b) Data visualizations

#### 2.b.i) Univariate plots to better understand each attribute.

```{r}
# Set up a variable for the total number of attribute columns (totAttr)
totAttr <- totCol-1
# Set up the number of row and columns for visualization display. dispRow * dispCol should be >= totAttr
dispCol <- 4
if (totAttr%%dispCol == 0) {
dispRow <- totAttr%/%dispCol
} else {
dispRow <- (totAttr%/%dispCol) + 1
}
```


```{r}
# Boxplots for each attribute
#par(mfrow=c(dispRow,dispCol))
#for(i in 1:totAttr) {
#	boxplot(entireDataset[,i], main=names(entireDataset)[i])
#}
```

```{r}
# Histograms each attribute
#par(mfrow=c(dispRow,dispCol))
#for(i in 1:totAttr) {
#	hist(entireDataset[,i], main=names(entireDataset)[i])
#}
```

```{r}
# Density plot for each attribute
#par(mfrow=c(dispRow,dispCol))
#for(i in 1:totAttr) {
#	plot(density(entireDataset[,i]), main=names(entireDataset)[i])
#}
```

#### 2.b.ii) Multivariate plots to better understand the relationships between attributes

```{r}
# Scatterplot matrix colored by class
#pairs(targetVar~., data=entireDataset, col=training$targetVar)
```

```{r}
# Box and whisker plots for each attribute by class
#scales <- list(x=list(relation="free"), y=list(relation="free"))
#featurePlot(x=entireDataset_x, y=entireDataset_y, plot="box", scales=scales)
```

```{r}
# Density plots for each attribute by class value
#featurePlot(x=entireDataset_x, y=entireDataset_y, plot="density", scales=scales)
```

```{r}
# Correlation plot
#correlations <- cor(entireDataset_x)
#corrplot(correlations, method="circle")
```

## 3. Prepare Data
Some training may require additional preparation activities that will best exposes the structure of the problem and the relationships between the input attributes and the output variable. Some data-prep tasks might include:

* Cleaning data by removing duplicates, marking missing values and even imputing missing values.
* Feature selection where redundant features may be removed.
* Data transforms where attributes are scaled or redistributed in order to best expose the structure of the problem later to learning algorithms.

### 3.a) Data Cleaning

```{r}
# Not applicable for this iteration of the project.

# Mark missing values
#invalid <- 0
#entireDataset$some_col[entireDataset$some_col==invalid] <- NA

# Impute missing values
#entireDataset$some_col <- with(entireDataset, impute(some_col, mean))
```

### 3.b) Feature Selection

```{r}
# Not applicable for this iteration of the project.
```

### 3.c) Data Transforms

```{r}
# Not applicable for this iteration of the project.
```

### 3.d) Split-out training and validation trainings
We create a training training (variable name "training") and a validation training (variable name "validation").

```{r}
seedNum <- 888
set.seed(seedNum)

# Create a list of 80% of the rows in the original training we can use for training
training_index <- createDataPartition(entireDataset$targetVar, p=0.80, list=FALSE)

# Use 80% of data to training and testing the models
training <- entireDataset[training_index,]

# Select the remaining 20% of the data for validation
validation <- entireDataset[-training_index,]
```

## 4. Model and Evaluate Algorithms
After the data-prep, we next work on finding a workable model by evaluating a subset of machine learning algorithms that are good at exploiting the structure of the training. The typical evaluation tasks include:

* Defining test options such as cross validation and the evaluation metric to use.
* Spot checking a suite of linear and nonlinear machine learning algorithms.
* Comparing the estimated accuracy of algorithms.

For this project, we will evaluate two linear, three non-linear, and four ensemble algorithms:

Linear Algorithms: Logistic Regression and Linear Discriminant Analysis

Non-Linear Algorithms: Decision Trees (CART), k-Nearest Neighbors, and Support Vector Machine

Ensemble Algorithms: Bagged CART, Random Forest, ADA, and Stochastic Gradient Boosting

The random number seed is reset before each run to ensure that the evaluation of each algorithm is performed using the same data splits. It ensures the results are directly comparable.

### 4.a) Set test options and evaluation metric

```{r}
# Run algorithms using 10-fold cross validation
control <- trainControl(method="repeatedcv", number=10, repeats=3)
metricTarget <- "Accuracy"
```

### 4.b) Generate models using linear algorithms

```{r LR}
# Logistic Regression (Classification)
set.seed(seedNum)
fit.glm <- train(targetVar~., data=training, method="glm", metric=metricTarget, trControl=control)
```

```{r LDA}
# Linear Discriminant Analysis (Classification)
set.seed(seedNum)
fit.lda <- train(targetVar~., data=training, method="lda", metric=metricTarget, trControl=control)
```

### 4.c) Generate models using nonlinear algorithms

```{r CART}
# Decision Tree - CART (Regression/Classification)
set.seed(seedNum)
fit.cart <- train(targetVar~., data=training, method="rpart", metric=metricTarget, trControl=control)
```

```{r KNN}
# k-Nearest Neighbors (Regression/Classification)
set.seed(seedNum)
fit.knn <- train(targetVar~., data=training, method="knn", metric=metricTarget, trControl=control)
```

```{r SVM}
# Support Vector Machine (Classification)
set.seed(seedNum)
fit.svm <- train(targetVar~., data=training, method="svmRadial", metric=metricTarget, trControl=control)
```

### 4.d) Generate models using ensemble algorithms
In this section, we will explore the use and tuning of ensemble algorithms to see whether we can improve the results.

```{r BAGCART}
# Bagged CART (Regression/Classification)
set.seed(seedNum)
fit.bagcart <- train(targetVar~., data=training, method="treebag", metric=metricTarget, trControl=control)
```

```{r RF}
# Random Forest (Regression/Classification)
set.seed(seedNum)
fit.rf <- train(targetVar~., data=training, method="rf", metric=metricTarget, trControl=control)
```

```{r GBM}
# Stochastic Gradient Boosting (Regression/Classification)
set.seed(seedNum)
fit.gbm <- train(targetVar~., data=training, method="gbm", metric=metricTarget, trControl=control)
```

### 4.e) Compare baseline algorithms

```{r SPOT_CHECK}
results <- resamples(list(LogReg=fit.glm, LDA=fit.lda, CART=fit.cart, kNN=fit.knn, SVM=fit.svm, BagCART=fit.bagcart, RF=fit.rf, GBM=fit.gbm))
summary(results)
dotplot(results)
```

## 5. Improve Accuracy or Results
After we achieve a short list of machine learning algorithms with good level of accuracy, we can leverage ways to improve the accuracy of the models:

* Search for a combination of parameters for each algorithm using caret that yields the best results.
* Combine the prediction of multiple models into an ensemble prediction using standalone algorithms or the caret Ensemble package.

### 5.a) Algorithm Tuning
Finally, we will tune the best-performing algorithms from each group further and see whether we can get more accuracy out of them.

```{r FINAL1}
# Tuning algorithm #1 - Bagged CART
# Bagged CART requires no special tuning parameters.
set.seed(seedNum)
fit.final1 <- fit.bagcart
print(fit.final1)
```

```{r FINAL2}
# Tuning algorithm #2 - Random Forest
set.seed(seedNum)
grid <- expand.grid(mtry = c(1:10))
fit.final2 <- train(targetVar~., data=training, method="rf", metric=metricTarget, tuneGrid=grid, trControl=control)
plot(fit.final2)
print(fit.final2)
```

```{r FINAL3}
# Tuning algorithm #3 - Stochastic Gradient Boosting
set.seed(seedNum)
grid <- expand.grid(.n.trees=c(100, 250, 500), .interaction.depth=c(1, 2, 3), .shrinkage=c(0.001, 0.025, 0.05, 0.1), .n.minobsinnode=c(5, 10, 15))
fit.final3 <- train(targetVar~., data=training, method="gbm", metric=metricTarget, tuneGrid=grid, trControl=control)
plot(fit.final3)
print(fit.final3)
```

### 5.d) Compare Algorithms After Tuning

```{r POST_TUNING}
results <- resamples(list(BagCART=fit.final1, RF=fit.final2, GBM=fit.final3))
summary(results)
dotplot(results)
```

## 6. Finalize Model and Present Results
Once we have narrow down to a model that we believe can make accurate predictions on unseen data, we are ready to finalize it. Finalizing a model may involve sub-tasks such as:

* Using an optimal model tuned to make predictions on unseen data.
* Creating a standalone model using the tuned parameters
* Saving an optimal model to file for later use.

### 6.a) Predictions on validation training

```{r PREDICT}
predictClass <- predict(fit.final2, newdata=validation)
confusionMatrix(predictClass, validation$targetVar)

pred <- prediction(as.numeric(predictClass), as.numeric(validation$targetVar))
perf <- performance(pred, measure = "tpr", x.measure = "fpr")
plot(perf, colorize=TRUE)

auc <- performance(pred, measure = "auc")
auc <- auc@y.values[[1]]
auc
```

### 6.b) Create standalone model on entire training training

```{r}
library(randomForest)
set.seed(seedNum)
finalModel <- randomForest(targetVar~., entireDataset, mtry=10, na.action=na.omit)
summary(finalModel)
```

### 6.c) Save model for later use

```{r}
#saveRDS(finalModel, "./finalModel_BinaryClass.rds")
```
