---
title: "Binary Classification Model for Truck APS Failure Detection Using R Take 1"
author: "David Lowe"
date: "November 5, 2018"
output: 
  html_document: 
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Template Credit: Adapted from a template made available by Dr. Jason Brownlee of Machine Learning Mastery. https://machinelearningmastery.com/

SUMMARY: The purpose of this project is to construct a prediction model using various machine learning algorithms and to document the end-to-end steps using a template. The APS Failure at Scania Trucks dataset is a binary-class classification situation where we are trying to predict one of the two possible outcomes.

INTRODUCTION: The dataset consists of data collected from heavy Scania trucks in everyday usage. The system in focus is the Air Pressure system (APS) which generates pressurised air that is utilized in various functions in a truck, such as braking and gear changes. The datasets' positive class consists of component failures for a specific component of the APS system. The negative class consists of trucks with failures for components not related to the APS. The data consists of a subset of all available data, selected by experts.

This dataset has many cells with missing values, so it is not practical to simply delete the rows with missing cells. This iteration of the project will produce a set of baseline results by imputing the blank cells with the value zero.

ANALYSIS: The baseline performance of the ten algorithms achieved an average accuracy of 99.02%. Three ensemble algorithms (Bagged CART, Random Forest, and Stochastic Gradient Boosting) chieved the top accuracy scores after the first round of modeling. After a series of tuning trials, Random Forest turned in the top result using the training data. It achieved an average accuracy of 99.39%. Using the optimized tuning parameter available, the Random Forest algorithm processed the validation dataset with an accuracy of 99.22%, which was slightly below the accuracy of the training data.

CONCLUSION: For this iteration, the Random Forest algorithm achieved the top-tier training and validation results. For this dataset, the Random Forest algorithm should be considered for further modeling or production use.

Dataset Used: APS Failure at Scania Trucks Data Set

Dataset ML Model: Binary classification with numerical attributes

Dataset Reference: https://archive.ics.uci.edu/ml/datasets/APS+Failure+at+Scania+Trucks

One potential source of performance benchmark: https://www.kaggle.com/uciml/aps-failure-at-scania-trucks-data-set/home

The project aims to touch on the following areas:

1. Document a predictive modeling problem end-to-end.
2. Explore data cleaning and transformation options
3. Explore non-ensemble and ensemble algorithms for baseline model performance
4. Explore algorithm tuning techniques for improving model performance

Any predictive modeling machine learning project genrally can be broken down into about six major tasks:

1. Prepare Problem
2. Summarize Data
3. Prepare Data
4. Model and Evaluate Algorithms
5. Improve Accuracy or Results
6. Finalize Model and Present Results

## 1. Prepare Problem

### 1.a) Load libraries

```{r}
startTimeScript <- proc.time()
library(caret)
library(corrplot)
library(ROCR)
library(parallel)

# Create one random seed number for reproducible results
seedNum <- 888
set.seed(seedNum)
```

### 1.b) Load dataset

```{r}
xy_train <- read.csv("aps_failure_training_set.csv", header= TRUE, na.strings="na", stringsAsFactors = T)
xy_test <- read.csv("aps_failure_test_set.csv", header= TRUE, na.strings="na", stringsAsFactors = T)
```

```{r}
# Use variable totCol to hold the number of columns in the dataframe
totCol <- ncol(xy_train)

# Set up variable totAttr for the total number of attribute columns
totAttr <- totCol-1
```

```{r}
# targetCol variable indicates the column location of the target/class variable
# If the first column, set targetCol to 1. If the last column, set targetCol to totCol
# if (targetCol <> 1) and (targetCol <> totCol), be aware when slicing up the dataframes for visualization!
targetCol <- 1
colnames(xy_train)[targetCol] <- "targetVar"
colnames(xy_test)[targetCol] <- "targetVar"
```

```{r}
# We create training datasets (xy_train, x_train, y_train) for various operations.
# We create validation datasets (xy_test, x_test, y_test) for various operations.
set.seed(seedNum)

# Create a list of the rows in the original dataset we can use for training
# training_index <- createDataPartition(originalDataset$targetVar, p=0.70, list=FALSE)
# Use 70% of the data to train the models and the remaining test/validation
# xy_train <- originalDataset[training_index,]
# xy_test <- originalDataset[-training_index,]

if (targetCol==1) {
x_train <- xy_train[,(targetCol+1):totCol]
y_train <- xy_train[,targetCol]
y_test <- xy_test[,targetCol]
} else {
x_train <- xy_train[,1:(totAttr)]
y_train <- xy_train[,totCol]
y_test <- xy_test[,totCol]
}
```

### 1.c) Set up the key parameters to be used in the script

```{r}
# Set up the number of row and columns for visualization display. dispRow * dispCol should be >= totAttr
dispCol <- 5
if (totAttr%%dispCol == 0) {
dispRow <- totAttr%/%dispCol
} else {
dispRow <- (totAttr%/%dispCol) + 1
}
cat("Will attempt to create graphics grid (col x row): ", dispCol, ' by ', dispRow)
```

```{r}
# Run algorithms using 10-fold cross validation
control <- trainControl(method="repeatedcv", number=10, repeats=1)
metricTarget <- "Accuracy"
```

## 2. Summarize Data
To gain a better understanding of the data that we have on-hand, we will leverage a number of descriptive statistics and data visualization techniques. The plan is to use the results to consider new questions, review assumptions, and validate hypotheses that we can investigate later with specialized models.

### 2.a) Descriptive statistics

#### 2.a.i) Peek at the data itself.

```{r}
head(xy_train)
```

#### 2.a.ii) Dimensions of the dataset.

```{r}
dim(xy_train)
dim(xy_test)
```

#### 2.a.iii) Types of the attributes.

```{r}
sapply(xy_train, class)
```

#### 2.a.iv) Statistical summary of all attributes.

```{r}
summary(xy_train)
```

#### 2.a.v) Summarize the levels of the class attribute.

```{r}
cbind(freq=table(y_train), percentage=prop.table(table(y_train))*100)
```

#### 2.a.vi) Count missing values.

```{r}
sapply(xy_train, function(x) sum(is.na(x)))
```

### 2.b) Data visualizations

#### 2.b.i) Univariate plots to better understand each attribute.

```{r}
# Boxplots for each attribute
#par(mfrow=c(dispRow,dispCol))
#for(i in 1:totAttr) {
#	boxplot(x_train[,i], main=names(x_train)[i])
#}
```

```{r}
# Histograms each attribute
#par(mfrow=c(dispRow,dispCol))
#for(i in 1:totAttr) {
#	hist(x_train[,i], main=names(x_train)[i])
#}
```

```{r}
# Density plot for each attribute
#par(mfrow=c(dispRow,dispCol))
#for(i in 1:totAttr) {
#	plot(density(x_train[,i]), main=names(x_train)[i])
#}
```

#### 2.b.ii) Multivariate plots to better understand the relationships between attributes

```{r}
# Scatterplot matrix colored by class
#pairs(targetVar~., data=xy_train, col=xy_train$targetVar)
```

```{r}
# Box and whisker plots for each attribute by class
#scales <- list(x=list(relation="free"), y=list(relation="free"))
#featurePlot(x=x_train, y=y_train, plot="box", scales=scales)
```

```{r}
# Density plots for each attribute by class value
#featurePlot(x=x_train, y=y_train, plot="density", scales=scales)
```

```{r}
# Correlation plot
#correlations <- cor(x_train)
#corrplot(correlations, method="circle")
```

## 3. Prepare Data
Some dataset may require additional preparation activities that will best exposes the structure of the problem and the relationships between the input attributes and the output variable. Some data-prep tasks might include:

* Cleaning data by removing duplicates, marking missing values and even imputing missing values.
* Feature selection where redundant features may be removed.
* Data transforms where attributes are scaled or redistributed in order to best expose the structure of the problem later to learning algorithms.

### 3.a) Data Cleaning

```{r}
# Impute missing cells with the value zero for the training dataset
xy_train[is.na(xy_train)] <- 0
sapply(xy_train, function(x) sum(is.na(x)))
```

```{r}
# Impute missing cells with the value zero for the test dataset
xy_test[is.na(xy_test)] <- 0
sapply(xy_test, function(x) sum(is.na(x)))
```

### 3.b) Feature Selection

```{r}
# Not applicable for this iteration of the project.
```

### 3.c) Data Transforms

```{r}
# Not applicable for this iteration of the project.
```

```{r}
proc.time()-startTimeScript
```

## 4. Model and Evaluate Algorithms
After the data-prep, we next work on finding a workable model by evaluating a subset of machine learning algorithms that are good at exploiting the structure of the training. The typical evaluation tasks include:

* Defining test options such as cross validation and the evaluation metric to use.
* Spot checking a suite of linear and nonlinear machine learning algorithms.
* Comparing the estimated accuracy of algorithms.

For this project, we will evaluate one linear, three non-linear, and three ensemble algorithms:

Linear Algorithm: Linear Discriminant Analysis

Non-Linear Algorithms: Decision Trees (CART), k-Nearest Neighbors, and Support Vector Machine

Ensemble Algorithms: Bagged CART, Random Forest, and Stochastic Gradient Boosting

The random number seed is reset before each run to ensure that the evaluation of each algorithm is performed using the same data splits. It ensures the results are directly comparable.

### 4.a) Generate models using linear algorithms

```{r LDA}
# Logistic Regression (Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.glm <- train(targetVar~., data=xy_train, method="glm", metric=metricTarget, trControl=control)
print(fit.glm)
proc.time()-startTimeModule
```

### 4.b) Generate models using nonlinear algorithms

```{r CART}
# Decision Tree - CART (Regression/Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.cart <- train(targetVar~., data=xy_train, method="rpart", metric=metricTarget, trControl=control)
print(fit.cart)
proc.time()-startTimeModule
```

```{r KNN}
# k-Nearest Neighbors (Regression/Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.knn <- train(targetVar~., data=xy_train, method="knn", metric=metricTarget, trControl=control)
print(fit.knn)
proc.time()-startTimeModule
```

```{r SVM}
# Support Vector Machine (Regression/Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.svm <- train(targetVar~., data=xy_train, method="svmRadial", metric=metricTarget, trControl=control)
print(fit.svm)
proc.time()-startTimeModule
```

### 4.c) Generate models using ensemble algorithms
In this section, we will explore the use and tuning of ensemble algorithms to see whether we can improve the results.

```{r BAGCART}
# Bagged CART (Regression/Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.bagcart <- train(targetVar~., data=xy_train, method="treebag", metric=metricTarget, trControl=control)
print(fit.bagcart)
proc.time()-startTimeModule
```

```{r RF}
# Random Forest (Regression/Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.rf <- train(targetVar~., data=xy_train, method="rf", metric=metricTarget, trControl=control)
print(fit.rf)
proc.time()-startTimeModule
```

```{r GBM}
# Stochastic Gradient Boosting (Regression/Classification)
startTimeModule <- proc.time()
set.seed(seedNum)
fit.gbm <- train(targetVar~., data=xy_train, method="gbm", metric=metricTarget, trControl=control, verbose=F)
print(fit.gbm)
proc.time()-startTimeModule
```

### 4.d) Compare baseline algorithms

```{r SPOT_CHECK}
results <- resamples(list(LR=fit.glm, CART=fit.cart, kNN=fit.knn, SVM=fit.svm, BagCART=fit.bagcart, RF=fit.rf, GBM=fit.gbm))
summary(results)
dotplot(results)
cat('The average accuracy from all models is:',
    mean(c(results$values$`LR~Accuracy`,results$values$`CART~Accuracy`,results$values$`kNN~Accuracy`,results$values$`SVM~Accuracy`,results$values$`BagCART~Accuracy`,results$values$`RF~Accuracy`,results$values$`GBM~Accuracy`)))
```

## 5. Improve Accuracy or Results
After we achieve a short list of machine learning algorithms with good level of accuracy, we can leverage ways to improve the accuracy of the models.

Using the top two best-perfoming algorithms from the previous section, we will Search for a combination of parameters for each algorithm that yields the best results.

### 5.a) Algorithm Tuning
Finally, we will tune the best-performing algorithms from each group further and see whether we can get more accuracy out of them.

```{r FINAL1}
# Tuning algorithm #1 - Bagged CART
startTimeModule <- proc.time()
set.seed(seedNum)
fit.final1 <- fit.bagcart
print(fit.final1)
proc.time()-startTimeModule
```

```{r FINAL2}
# Tuning algorithm #2 - Random Forest
startTimeModule <- proc.time()
set.seed(seedNum)
grid <- expand.grid(mtry = c(5:9))
fit.final2 <- train(targetVar~., data=xy_train, method="rf", metric=metricTarget, tuneGrid=grid, trControl=control)
plot(fit.final2)
print(fit.final2)
proc.time()-startTimeModule
```

```{r FINAL3}
# Tuning algorithm #3 - Stochastic Gradient Boosting
startTimeModule <- proc.time()
set.seed(seedNum)
grid <- expand.grid(.n.trees=c(800,900,1000,1100,1200), .shrinkage=c(0.1), .interaction.depth=c(1), .n.minobsinnode=c(10))
fit.final3 <- train(targetVar~., data=xy_train, method="gbm", metric=metricTarget, tuneGrid=grid, trControl=control, verbose=F)
plot(fit.final3)
print(fit.final3)
proc.time()-startTimeModule
```

### 5.d) Compare Algorithms After Tuning

```{r POST_TUNING}
results <- resamples(list(BagCART=fit.final1, RF=fit.final2, GBM=fit.final3))
summary(results)
dotplot(results)
```

## 6. Finalize Model and Present Results
Once we have narrow down to a model that we believe can make accurate predictions on unseen data, we are ready to finalize it. Finalizing a model may involve sub-tasks such as:

* Using an optimal model tuned to make predictions on unseen data.
* Creating a standalone model using the tuned parameters
* Saving an optimal model to file for later use.

### 6.a) Predictions on validation dataset

```{r PREDICT}
predictions <- predict(fit.final2, newdata=xy_test)
confusionMatrix(predictions, xy_test$targetVar)
```

### 6.b) Create standalone model on entire training dataset

```{r}
startTimeModule <- proc.time()
library(randomForest)
set.seed(seedNum)
finalModel <- randomForest(targetVar~., data=xy_train, mtry=9)
summary(finalModel)
proc.time()-startTimeModule
```

### 6.c) Save model for later use

```{r}
#saveRDS(finalModel, "./finalModel_MultiClass.rds")
```

```{r}
proc.time()-startTimeScript
```
